---
title: "Impacts of Climate Change on Mortality: An extrapolation of temperature 
effects based on time series data in France"
author: "Quentin Guibert"
date: "2024-07-15"
header-includes:
  - \usepackage{amsmath,amsthm,amssymb,mathabx, bm, bbm, color}
  - \usepackage{dsfont}
  - \usepackage{stackengine}
  - \usepackage{scalerel}
  - \newcommand{\E}{\mathbb{E}}
  - \newcommand{\Yvec}{\boldsymbol{Y}}
  - \newcommand{\Avec}{\boldsymbol{A}}
  - \newcommand{\Evec}{\boldsymbol{E}}
  - \newcommand{\uvec}{\boldsymbol{u}}
  - \newcommand{\avec}{\boldsymbol{a}}
  - \newcommand{\bvec}{\boldsymbol{b}}
  - \newcommand{\nvec}{\boldsymbol{n}}
  - \newcommand{\rvec}{\boldsymbol{r}}
  - \newcommand{\mvec}{\boldsymbol{m}}
  - \newcommand{\cvec}{\boldsymbol{c}}
  - \newcommand{\gvec}{\boldsymbol{g}}
  - \newcommand{\hvec}{\boldsymbol{h}}
  - \newcommand{\pvec}{\boldsymbol{p}}
  - \newcommand{\dvec}{\boldsymbol{D}}
  - \newcommand{\Thetavec}{\boldsymbol{\Theta}}
  - \newcommand{\betavec}{\boldsymbol{\beta}}
  - \newcommand{\etavec}{\boldsymbol{\eta}}
  - \newcommand{\xivec}{\boldsymbol{\xi}}
  - \newcommand{\alphavec}{\boldsymbol{\alpha}}
  - \newcommand{\omegavec}{\boldsymbol{\omega}}
  - \newcommand{\Omegavec}{\boldsymbol{\Omega}}
  - \newcommand{\muvec}{\boldsymbol{\mu}}
  - \newcommand{\thetavec}{\boldsymbol{\theta}}
  - \newcommand{\lambdavec}{\boldsymbol{\lambda}}
  - \newcommand{\Lambdavec}{\boldsymbol{\Lambda}}
  - \newcommand{\Sigmavec}{\boldsymbol{\Sigma}}
  - \newcommand{\zetavec}{\boldsymbol{\zeta}}
  - \newcommand{\kappavec}{\boldsymbol{\kappa}}
  - \newcommand{\epsilonvec}{\boldsymbol{\epsilon}}
  - \newcommand{\gammavec}{\boldsymbol{\gamma}}
  - \newcommand{\deltavec}{\boldsymbol{\delta}}
  - \newcommand{\pivec}{\boldsymbol{\pi}}
  - \newcommand{\Pivec}{\boldsymbol{\Pi}}
  - \newcommand{\psivec}{\boldsymbol{\psi}}
  - \newcommand{\Psivec}{\boldsymbol{\Psi}}
  - \newcommand{\Gammavec}{\boldsymbol{\Gamma}}
  - \newcommand{\phivec}{\boldsymbol{\phi}}
  - \newcommand{\Phivec}{\boldsymbol{\Phi}}
  - \newcommand{\varphivec}{\boldsymbol{\varphi}}
  - \newcommand{\evec}{\boldsymbol{e}}
  - \newcommand{\ovec}{\boldsymbol{o}}
  - \newcommand{\fvec}{\boldsymbol{f}}
  - \newcommand{\Hvec}{\boldsymbol{H}}
  - \newcommand{\Qvec}{\boldsymbol{Q}}
  - \newcommand{\qvec}{\boldsymbol{q}}
output:
  rmdformats::robobook:
    highlight: kate
    use_bookdown: yes
    css: style.css
    lightbox: yes
    gallery: yes
    code_folding: show
  code-copy: yes
  html-math-method: katex
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
institute: "CEREMADE, Université Paris Dauphine-PLS"
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

```{=html}
<style>
body { text-align: justify}
</style>
```
This R Markdown Notebook reproduces the numerical applications presented
in the paper "Impacts of Climate Change on Mortality: An extrapolation
of temperature effects based on time series data in France", available
[here](https://arxiv.org/abs/2406.02054).

The objective of this Notebook is to illustrate the estimation techniques and the
presentation of results described in this article. Since some mortality data are
not publicly available, the estimates obtained from the DLNM model are presented
but cannot be reproduced. However, we present the functions to be applied,
enabling the reader to replicate the process with their own data.

For the sake of brevity and code clarity, we limit the presentation to the
aggregated results in Metropolitan France only. We also consider only the
scenarios related to one climate model instead of the 12 presented in this paper.
This simplification does not limit the understanding of the code or the
reproducibility of the results.

Note that we use an adjusted version of the
[`MultiMoMo`](https://github.com/jensrobben/MultiMoMo) package. Some functions
have been recoded or added to better suit our application.

# Set-up

First, we set up the working environment and load the necessary
functions and packages for the application.

```{r setup, include=FALSE}
# Global options
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, 
                      fig.retina = 2)
options(encoding = 'UTF-8')
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Folders
fold <- getwd()
fold_data <- paste0(fold, "/data/")
fold_bib <- paste0(fold, "/functions/")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Load functions
for (f in list.files(fold_bib))
  source(paste(fold_bib,f,sep=""), encoding = "UTF-8")
#----------------------------------------------------------
# Load packages
invisible(lapply(c(
  "MultiMoMo",
  "HMDHFDplus",
  "systemfit",
  "lattice",
  "grid",
  "ggplot2",
  "gridExtra",
  "locfit",
  "scales",
  "lubridate",
  "splines",
  "mgcv",
  "dlnm",
  "MASS",
  "mvtnorm",
  "Rfast",
  "parallel",
  "data.table",
  "formattable",
  "RColorBrewer",
  "readxl",
  "kableExtra",
  "dplyr",
  "tidyr",
  "tidyverse",
  "rmarkdown",
  "ggthemes",
  "cowplot",
  "ggridges",
  "viridis",
  "hrbrthemes",
  "colorspace",
  "ggbeeswarm",
  "ggpubr"),
  instal.import.package))
options(dplyr.summarise.inform = FALSE)
#----------------------------------------------------------
# Graphical parameters
theme_set(theme_bw())
trellis.device(color = FALSE)
```

```{r, echo = TRUE, message=FALSE, warning=FALSE}


# Forecasting period
start_year <- 2020
end_year <- 2100
year_breaks <- c(1980, 1989, 1999, 2009, 2019, 2029, 2039, 2049, 2059, 
                 2069, 2079, 2089, 2100)
year_labels <- c("1980s", "1990s", "2000s", "2010s", "2020s", "2030s", 
                 "2040s", "2050s", "2060s", "2070s", "2080s","2090s")

# Heatwave parameter - french definition
min_level <- 18
max_level <- 30

# Monte-Carlo simulations
nsim <- 20 # number of simulations
parallel <- "no" # parallel option
ncpus <- 1L # number of cores

# Other parameters
AGE_MAX <- 105
```

# Load data

In this section, we start by loading the input data. Three types of information
are considered: historical mortality data (including daily temperatures and
daily deaths), annual mortality data, and climate scenarios.

More information regarding the data description is available in [the
paper](https://arxiv.org/abs/2406.02054).

## Daily Mortality data

We import daily data series for death counts (which are not freely available)
and temperatures. This dataset is constructed by merging:

-   daily mortality data,
-   daily temperature data.

```{r, eval=TRUE, message=FALSE, warning=FALSE}
daily_data <- get(load(file =  paste0(fold_data, "/daily_data.RData")))

# Summary of the data - list with 2 items for females (`f`) and males (`m`)
head(daily_data)

# Get quantiles for extreme hot and cold of average daily temperature
q025 <- quantile(daily_data$m$tavg, 0.025)
q975 <- quantile(daily_data$m$tavg, 0.975)
```

The object `daily_data` is a list containing two datasets for females (`f`) and
males (`m`). Each dataset contains:

-   `datedec`: the date,
-   `age_bk`: age buckets,
-   `nb_deaths`: number of deaths,
-   `years`: years,
-   `tmax`: maximum temperature of the day
-   `tmax3`: maximum temperature over the last 3 days
-   `tmin`: minimum temperature over the last 3 days
-   `tmin3`: minimum temperature over the last 3 days
-   `tavg`: average temperature of the day
-   `tavg3`: average temperature over the last 3 days
-   `time`: day number as a integer number
-   `dow`: day of the week as a integer number
-   `month`: month number

Our DLNM model for `nb_deaths` is fitted based on the `tavg` variable, as
well as `datedec`, `dow` and `month`.

## Annual mortality data from the HMD

The annual mortality data considered are extracted from the Human Mortality 
Database \citep{human_mortality_database_university_2024}. We downloaded the
exposure to risk and death counts from the
[HMD website](https://www.mortality.org/). We consider data from Metropolitan
France for the calibration period $\mathcal{T}_y = \{1980, \ldots, 2019\}$
and the age range $\mathcal{X} = \{0, \ldots, 105\}$.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
mort_dt <- get(load(file =  paste0(fold_data, "/mort_dt.RData")))

```


## Data from climate models

We import daily data series for climate trajectories. The models selected in the
paper are the 12 models available through the \citet{drias} portal for
RCP2.6, RCP4.5, and RCP8.5 scenarios. In this Notebook, we only consider
the model CNRM-CM5 / ALADIN63 with the RCP8.5 scenario.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
clim_model <- get(load(file = paste0(fold_data, 
                                     "ALADIN63_CNRM-CM5_RCP8.5.RData")))
```

# Estimating the DLNM

The distributed lag non-linear generalized model (DLNM) is a gold standard in 
climate epidemiology for assessing the impact of delayed environmental 
factors on a response variable [@gasparrini_distributed_2010, @guo_gasparrini14].

We fit DLNM models for each gender and each age bucket. This model aims to assess
the influence of temperature on the daily number of deaths. The number of daily
deaths $D_{k,t,d}^{(g)}$, aggregated by $K$ age groups and by sex, for each day
$d \in D^\star = \left\lbrace 1,2,\ldots, 365, (366)\right\rbrace$
of year $t$ is modeled by
$$
\ln( \E [D_{k,t,d}^{(g)}])=\eta_k + s(\vartheta_{d,t}, L; \thetavec_k ) 
+ \uvec_d \gammavec_k^\top
+ \sum_{p=1}^{P}{h_p( z_{d,p} ; \zetavec_k )},
$$
where: 

  - $s (\vartheta_{d,t} ; l, \beta_{k})$ is a cross-basis (non-linear) function,
  capturing the cumulated effect of the daily mean temperature $\vartheta_{d,t}$
  over a maximum of $L$ days,
  - $h_p(z_{d,p} ; \zetavec_k )$ are natural cubic spline to account for seasonal
  variations, e.g year, day of the week or month as features.
  - confounding variables $\uvec_d$ could be integrated as a linear effect,
  e.g. pollutant. 

We consider a bi-dimensional spline function $s$ as the surface of Relative Risk 
(RR)
$$
s(\vartheta_{d,t} ; L, \thetavec_k ) = \int_{0}^{L}{f\cdot w(\vartheta_{d-l,t},l;\thetavec_k)dl} \approx \sum_{l=0}^{L}{f\cdot w(\vartheta_{d-l,t},l;\thetavec_k)},
$$
where $f\cdot w$ is a bi-dimensional integrable function, and $\thetavec_{k}$ a
vector of parameters. 

All models are calibrated based on daily temperature data over the
period 1980-2019, using the **dlnm** package [@dlnm14]. Cubic spline with internal
knots placed at the 10th, 75th, and 90th percentiles of the daily
temperature distribution is considered. The lag $L$ is equal to 21 days.

## Parameters

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F}

# Segmentation parameters
list_sexe <- list(m = 1, f = 2)
age_breaks <- c(0, 64, 74, 84, Inf)
age_labels <- c("0-64", "65-74", "75-84", "85+")

# Training period
annee_deb <- 1980
annee_fin <- 2019

# DLNM Parameters
param_dlnm_whole <- list(
  # main model, cubic natural spline with three internal knots in
  # the 10th, 75th, 90th percentiles of the temperature distribution
  varfun = "bs",
  vardegree = 2,
  varper  = c(10, 75, 90),
  ## specification of the lag function
  # Definition of the maximum lag, that is, 21 days
  lag  = 21,
  lagnk  = 3,
  ## degree of freedom for seasonality
  dfseas  = 8,
  ## degree of freedom for trend
  dftrend = NULL
  )
```

## Model estimation

We fit the model for each sex and age bucket.

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F, comment= F}
fit <- lapply(daily_data, function(x)
{
  fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)
}
)
```

## Results and diagnostics

We calculate the historical count of deaths attributed to temperatures for each
day of the year $d \in \mathcal{T}_d$, and age $x \in X_k$, $k \in \{1, \ldots, K\}$ 
using the formula
$$
\bar{D}_{x,t,d}^{(g)}= (1-\exp{\left(- s(\vartheta_{d,t} ; L, \widehat{\thetavec}_k ) \right)}) \times D_{x,t,d}^{(g)}.
$$
The term $\exp{\left(  s(\vartheta_{d,t} ; L, \widehat{\thetavec}_k ) \right)}$ 
corresponds to the RR of temperature mortality. We display below the RR curves 
for temperatures.

```{r, echo = TRUE, fig.width=12, fig.height=8 , message=T, warning=T}
summary_dlnm_all_sex(fit)
```

We then analyze the goodness of fit performance of our models. For this, we 
display the residuals of the Deviance by year and age group, and compare the
distribution of observed (in blue) and predicted (in green) mortality counts for
each month, distinguishing by sex and decade.

```{r, echo = TRUE, fig.width=12, fig.height=8 , message=T, warning=T}
# in-sample performance of the fitted model for the whole year.
perf_mort <- lapply(names(daily_data), function(x){
  perf_dnlm(fit[[x]])
})
names(perf_mort) <- names(daily_data)
```

### Females

```{r, echo=FALSE, fig.width=12, fig.height=12 , message=F, warning=F}
perf_mort$f$residual
perf_mort$f$perf
```

### Males

```{r dlnm_resid_m, echo=FALSE, fig.width=12, fig.height=12 , message=F, warning=F}
perf_mort$m$residual
perf_mort$m$perf
```

## Estimation of the temperature- attributable fractions

The model's setup allows for a range of predictions that precisely gauge the 
influence of heat, cold, or specific events occurring on each days of subset 
$D \subseteq D^{\star}$.

The sum of the contributions from each day of this subperiod $D$
for a year $t \in \mathcal{T}_y$ is as follow
$$
\bar{D}_{x,t}^{(g)} = \sum_{d \in D}{\bar{D}_{x,t,d}^{(g)}\mathbf{1}_{\left\lbrace d \in D\right\rbrace }}.
$$

Thus, the  total attributable fraction is estimated for each age $x \in X_k$ and
year $t \in \mathcal{T}_y$ as 
$$
\theta_{x,t}^{(g)}= \dfrac{\bar{D}_{x,t}^{(g)}}{D_{x,t}^{(g)}}.
$$

We compute the temperature-attributable fractions $\theta^{(g)}_{x,t}$ for
both women and men, and aggregate them over all ages to facilitate
visualization on the figure below. The estimation error is calculated based 
on `r nsim` replications of parametric bootstraps.

```{r, echo=TRUE, fig.width=12, fig.height=6 , message=T, warning=T}

# Run excess of mortality
xs_mort <- lapply(names(daily_data), function(x){
  res <- excess_mort_dlnm(daily_data[[x]], fit[[x]], q_range = c(q025, q975),
                          nsim = nsim, parallel = parallel, ncpus = ncpus)
  return(res)
  })
names(xs_mort) <- names(daily_data)

# arrange data
xs_data <- lapply(names(xs_mort), function(i){
  xs_mort[[i]]$excess_mort %>%
    dplyr::mutate(gender = i) %>%
    rename(year = years)
})
xs_data <- do.call("rbind", xs_data)

# print attributable fractions
summary_temp_effect(xs_data)
```

# The Li-Lee model

We decomposition the number of deaths into two components
$$
D_{x,t}^{(g)} = \widetilde{D}_{x,t}^{(g)} + \bar{D}_{x,t}^{(g)} \Rightarrow 
\widehat{m}_{x,t}^{(g)} = \widetilde{m}_{x,t}^{(g)} + \bar{m}_{x,t}^{(g)}
$$
where $\widetilde{D}_{x,t}^{(g)}$ and $\bar{D}_{x,t}^{(g)}$ are respectively
the number of deaths without and with considering temperature effects.

Then, we consider the two-populations @li_coherent_2005 model for central deaths 
rates without temperature effects 
$$
\ln ~ \widetilde{m}_{x,t}^{\left(g\right)} = A_x + B_{x}K_{t} + \alpha_{x}^{\left( g \right)} + \beta_{x}^{\left( g \right)} \kappa_{t}^{\left( g \right)}.
$$		
We use the following specifications:

   - **Identifiability constraints**
   
$$
		\sum\limits_{t \in \mathcal{T}} K_{t} = 0 \text{ and } \sum\limits_{x \in \mathcal{X}} 
		B_{x}^2 = 1,
$$
$$
	  \sum\limits_{t \in \mathcal{T}_y} \kappa_{t}^{\left( g \right)} = 0  \text{ and } 
		\sum\limits_{x \in \mathcal{X}} (\beta_{x}^{\left( g \right)})^2 = 1,
		\text{ for } g \in \mathcal{G}.
$$

  - **Time series model with coherence assumption**
  
$$
K_{t} = \delta + K_{t-1} + e_{t} \;\; (\text{RWD with drift})
$$
$$
\kappa_{t}^{\left( g \right)} = c^{\left( g \right)} + \phi^{\left( g \right)}
\kappa_{t-1}^{\left( g \right)} + r_{t}^{\left( g \right)} \;\;
(\text{AR(1) with drift and } \vert \phi^{(g)} \vert <1).
$$
Errors terms are white noises with a mean of zero and a variance-covariance 
matrix $\Sigmavec$.

## Fitting the Poisson regression model 

For calibrating the parameters $({A}_x, {B}_x, {K}_{t},
{\alpha}_{x}^{(f)}, {\beta}_{x}^{(f)}, {\kappa}_{t}^{(f)},
{\alpha}_{x}^{(m)}, {\beta}_{x}^{(m)}, {\kappa}_{t}^{(m)})$, we use a Poisson
assumption for the number of virtual deaths as in @brouhns02

$$
\widetilde{D}_{x,t}^{(g)} \sim \text{Pois}\left( E_{x,t}^{(g)} \exp{\left( \widetilde{m}_{x,t}^{(g)}\right) }\right).
$$	
Thus,	with  the attributable fraction $\theta^{(g)}_{x,t}$, we also have a 
Poisson formulation with log-link function for the overall number of deaths
$$
{D}_{x,t}^{(g)} \sim \text{Pois}\left( E_{x,t}^{(g)} T_{x,t}^{(g)} \exp{\left( \widetilde{m}_{x,t}^{(g)}\right)}\right),
$$
where $T_{x,t}^ {(g)} = \exp{\left( \bar{m}_{x,t}^{(g)}\right)} = 1 + \theta^{(g)}_{x,t}$,
can be seen as an offset term.

### Calculating the offset term based on the attributable fraction

First, we define the mortality model hyper-parameters,i.e. the age range, the
observation period, and the name of each group.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
xv <- 0:105
yv <- 1980:2019
countries <- c("FRATNP")
group <- c("Female", "Male", "Total")
```

Then, we select `all_effect` (all the effects of temperatures over the year)
and assume that the offset term $T_{x,t}^{(g)}$ is constant for all ages
in each bucket.

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F}
# Data for women and men
xs_data <- get(load(file =  paste0(fold_data, "/xs_data.RData")))

xs_f <- xs_data %>% 
  dplyr::filter(gender == "f", temp_effect  == "all_effect") %>%
  dplyr::select(year, age_bk, attrib_frac)

xs_m <- xs_data %>% 
  dplyr::filter(gender == "m", temp_effect  == "all_effect") %>%
  dplyr::select(year, age_bk, attrib_frac)

# Transformation to matrix
transf_attrib_matrix <- function(df){
  mat <- matrix(data = 1, ncol = length(xv), nrow = length(yv))
  for(tt in yv){
    mat[tt - yv[1] + 1, 1:65] <- dplyr::filter(df, year == tt,
                                               age_bk == "0-64")$attrib_frac
    mat[tt - yv[1] + 1, 66:75] <- dplyr::filter(df, year == tt,
                                                age_bk == "65-74")$attrib_frac
    mat[tt - yv[1] + 1, 76:85] <- dplyr::filter(df, year == tt,
                                                age_bk == "75-84")$attrib_frac
    mat[tt - yv[1] + 1, 86:(max(xv) + 1)] <- dplyr::filter(
      df, year == tt, age_bk == "85+")$attrib_frac
  }
dimnames(mat) <- list(yv, xv)
return(mat)
}

xs_f <- transf_attrib_matrix(xs_f)
xs_m <- transf_attrib_matrix(xs_m)
```

### Adjusting exposure to risk

Now, we multiply the offset term $T_{x,t}^{(g)}$ by the exposures to risk. 
Subsequently, we construct a new dataset that we will use to fit the Li-Lee model.

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=T, warning=T}
# Create datasets for deaths with ajusted exposure
adj_mort_dt <- list(
  UNI = list(
    f = list(dtx = round(mort_dt$UNI$f$dtx),
             etx = mort_dt$UNI$f$etx * (1 + xs_f),
             wa = mort_dt$UNI$f$wa),
    m = list(dtx = round(mort_dt$UNI$m$dtx),
             etx = mort_dt$UNI$m$etx * (1 + xs_m),
             wa = mort_dt$UNI$m$wa)),
  ALL = list(dtx = mort_dt$ALL$dtx,
             etx = mort_dt$ALL$etx,
             wa = mort_dt$ALL$wa)
)
## replace exposure to adjusted exposure
adj_mort_dt$ALL$etx <- adj_mort_dt$UNI$f$etx + adj_mort_dt$UNI$m$etx
```

### Model calibration

We calibrate the parameters $({A}_x, {B}_x, {K}_{t},
{\alpha}_{x}^{(f)}, {\beta}_{x}^{(f)}, {\kappa}_{t}^{(f)},
{\alpha}_{x}^{(m)}, {\beta}_{x}^{(m)}, {\kappa}_{t}^{(m)})$
of the Li-Lee model using both observed and adjusted exposures to risk.
This allows to assess the impact of temperature-attributable deaths on these
estimated parameters.

These models are estimated by maximizing the Poisson log-likelihood. As 
detailed in the paper, our procedure consists in two steps. First, we estimate
the common parameters $({A}_x, {B}_x, {K}_{t})$,  followed by the estimation
of sex-specific parameters. This estimation is performed 
using the `R`-package **MultiMoMo** [@antonio_multimomo_2022].

```{r, eval = TRUE}
# Models with observed exposure to risk
fit_M <- fit_li_lee(xv, yv, yv, "m", data = mort_dt, method = "NR",
                    Ax = TRUE, exclude_coh = FALSE)
fit_F <- fit_li_lee(xv, yv, yv, "f", data = mort_dt, method = "NR",
                    Ax = TRUE, exclude_coh = FALSE)
```

```{r, eval = TRUE}
# Models with adjusted exposure to risk
adj_fit_M <- fit_li_lee(xv, yv, yv, "m", data = adj_mort_dt, method = "NR",
                        Ax = TRUE, exclude_coh = FALSE)
adj_fit_F <- fit_li_lee(xv, yv, yv, "f", data = adj_mort_dt, method = "NR",
                        Ax = TRUE, exclude_coh = FALSE)
```

The calibrated parameters are displayed as follow.

```{r, fig.width=12, fig.height=12 }
list_fit <- list(fit_F = fit_F,
                 fit_M = fit_M,
                 virt_fit_F = adj_fit_F,
                 virt_fit_M = adj_fit_M)
# print figures
myplot_parameters_li_lee(xv, yv, list_fit)
```

### Goodness of fit

We now display residuals for the model with exposures adjusted of temperature 
effects (model with original exposure provides similar results).

```{r, fig.width=12, fig.height=4}
fig_res_f <- plot_residuals_li_lee(xv, yv, adj_fit_F, "FR", "Female")
fig_res_m <- plot_residuals_li_lee(xv, yv, adj_fit_M, "FR", "Male")

plot_grid(fig_res_f + ggtitle("Females"), 
          fig_res_m + ggtitle('Males'),
          label_size = 12)
```

## Calibrating and forecasting the time series model 

In this section, we focus on calibrating the time series model, rewritten in matrix 
form as
$$
\Yvec_{t} = \dvec + \Phivec  \Yvec_{t-1} + \Evec_t,
$$
where
$$
\Yvec_{t} = \begin{pmatrix} K_t \\ \kappa_t^{(f)} \\ \kappa_t^{(m)} \end{pmatrix},  \dvec = \begin{pmatrix} \delta \\ c^{(f)} \\ c^{(m)} \end{pmatrix},
\Phivec = \begin{pmatrix}
1 & 0 & 0\\
0 & \phi^{(f)}  & 0\\
0 & 0  & \phi^{(m)}\\
\end{pmatrix} \text{ and }
\Evec_t = \begin{pmatrix} e_t \\ r_t^{(f)} \\ r_t^{(m)} \end{pmatrix}.
$$

We calibrate this model using two sets of parameters for $\Yvec_{t}$, i.e. 
the parameters obtained with the observed risk exposures and those obtained 
with the risk exposures adjusted for the effect of temperatures. These 
parameters are also estimated through maximum likelihood using `R`-package 
**MultiMoMo**. 

### Specification

We consider the projection period from the last year of the calibration set to
$H$, the projection horizon. 

```{r}
## Forecasting Parameters
arima_spec <- list(K.t_M = "RWD", k.t_M = "AR1.1", k.t_F = "AR1.1")
n_ahead    <- length(start_year:end_year)
n_sim      <- nsim
est_method <- "PORT"
```

### Calibration and projection

In this section, we estimate the parameters 
$\widehat{D}$, $\widehat{\Phi}$, and $\widehat{\Sigmavec}$. Then, we forecast
$\Yvec_t$ by simulating the innovation errors $\Evec_t$ from a multivariate
Gaussian vector with zero mean and covariance matrix $\widehat{\Sigmavec}$.
In this application, `r n_sim` Monte Carlo simulations is considered.

```{r}
set.seed(1234)
# Model with original exposures
proj_par <- project_parameters(fit_M, fit_F, n_ahead, n_sim,
                               arima_spec, est_method)
# Model with adjusted exposures
adj_proj_par <- project_parameters(adj_fit_M, adj_fit_F, n_ahead, n_sim,
                                   arima_spec, est_method)
```

We verify that the series modeled by an AR(1) process are stationary, i.e.
parameter estimates $\widehat{k}_t^{(g)}$ are smaller than 1.

```{r}
tab_param <- data.frame(
  v1 = c("Without temp. effects", "With temp. effects"),
  v2 = c(proj_par$coef_KtM, adj_proj_par$coef_KtM),
  v3 = c(proj_par$coef_ktM[1], adj_proj_par$coef_ktM[1]),
  v4 = c(proj_par$coef_ktM[2], adj_proj_par$coef_ktM[2]),
  v5 = c(proj_par$coef_ktF[1], adj_proj_par$coef_ktF[1]),
  v6 = c(proj_par$coef_ktF[2], adj_proj_par$coef_ktF[2])
)
# Print table
tab_param %>%
  kbl( booktabs = T,
       align='c',
       col.names = c("Calibration data", "$\\delta$",
                     "$c^{(m)}$", "$\\phi^{(m)}$",
                     "$c^{(f)}$", "$\\phi^{(f)}$"),
       digits=4, 
       escape =  FALSE) %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(latex_options = "hold_position",  font_size = 8)
```

The estimated covariance matrix $\boldsymbol{C}$ is given as follow.

```{r}
# Create correction matrix
corr_function <- function(S)
  {
    D <- diag(sqrt(diag(S)))
    Dinv <- solve(D)
    R <- Dinv %*% S %*% Dinv
    dimnames(R) <- dimnames(S)
    R 
  }

# Show correction matrix in table
tab_corr <- cbind(data.frame(corr_function(proj_par$cov_mat)), 
                  data.frame(corr_function(adj_proj_par$cov_mat)))
names(tab_corr) <- paste0("v", 1:ncol(tab_corr))
row.names(tab_corr) <- NULL

# Print table
tab_corr %>%
  dplyr::mutate(v = c("$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$"),
                .before = "v1") %>%
  kbl(booktabs = T,
       align='c',
       col.names = c("", "$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$",
                     "$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$"),
      digits=4, 
      escape =  FALSE) %>%
  add_header_above(c(" " = 1, "Without temp. effects" = 3,
                     "With temp. effects" = 3)) %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(latex_options = "hold_position",  font_size = 8)
```

### Projecting parameters

We present the projection of parameters $\widehat{K}_t$, $\widehat{\kappa}_t^{(f)}$,
and $\widehat{\kappa}_t^{(m)}$ over the period 2020-2100 with the $95\%$ prediction
intervals. These projections are shown for both models with and without 
temperature effects.

```{r, fig.width=12, fig.height=4}
p <- plot_kt_sim(proj_par, adj_proj_par, n_sim, yv[1], end_year)
print(p)
```

# Multi-mortality and climate models

Now that mortality rates excluding temperature effects can be projected,
we focus on adding mortality shocks resulting from daily temperature variations.
To achieve this, we consider a temperature trajectory derived from an external 
climate model $(\vartheta_d)$, and then calculate the daily number of deaths
attributed to temperature for each age $x$, sex $g$ and for any day $d$
of the year $t$ using
$$
\bar{D}_{x,t,d}^{(g)} = \widetilde{D}_{x,t,d}^{(g)}  \left( \exp{\left(s(\vartheta_{d,t} ; L, \widehat{\thetavec}_k ) \right)} - 1 \right).
$$
We deduce the cumulative number of deaths over a subperiod 
$D \subseteq D^{\star}$
$$
\bar{D}_{x,t}^{(g)} = \sum_{d \in D}{\bar{D}_{x,t,d}^{(g)}\mathbf{1}_{\left\lbrace d \in D\right\rbrace }}.
$$
The projected central mortality rates with temperature effects accumulated
over period $D$ are 
$$
\widehat{m}_{x,t}^{(g)} = 
 \widetilde{m}_{x,t}^{(g)} \left[ 1 + \overbrace{\sum_{d \in D}{\omega_{x,t,d}^{(g)} \left( \exp{\left(s(\vartheta_{d,t} ; L, \widehat{\thetavec}_k ) \right)} - 1 \right) \mathbf{1}_{\left\lbrace d \in D\right\rbrace }}}^{\theta_{x,t}^{(g)} (D)}\right],
$$
where $w_{x,t,d}^{(g)} =\widetilde{D}_{x,t,d}^{(g)}/\widetilde{D}_{x,t}^{(g)}$, a weight to be chosen, for the distribution of virtual deaths, e.g. $w_{x,t,d}^{(g)} = 1/D^\star$. 	

## Projecting future mortality

Now, we simulate the total mortality rates both with and without temperature effects
$$
q_{x,t}^{(g)} = 1 - \exp{\left(-\widehat{m}_{x,t}^{(g)}\right)},
\widetilde{q}_{x,t}^{(g)} = 1 - \exp{\left(-\widetilde{m}_{x,t}^{(g)}\right)}, 
$$

For each simulation, we also compute the number of years of life expectancy lost
(or gained) due to temperatures
$$
\Delta e_{x,t}^{(g)} = \int_{x}^{t_{\max}}{e^{- \int_{x}^{t}{\widetilde{\mu}_{x,u}^{(g)}du}}dt} 
- \int_{x}^{t_{\max}}{e^{- \int_{x}^{t}{\mu_{x,u}^{(g)}du}}dt} 
\approx \sum_{k=1}^{t_{\max}}{\left[
\prod_{j=0}^{k-1}{\left(1-\widetilde{q}_{x,j}^{(g)}\right)}
- \prod_{j=0}^{k-1}{\left(1-q_{x,j}^{(g)}\right)}
\right]},
$$
with the force of mortality
$\mu_{x,t}^{(g)} = \widetilde{\mu}_{x,t}^{(g)} + \bar{\mu}_{x,t}^{(g)}$.

The first step is to simulate the total death rates derived from models
calibrated on data, whether adjusted or not for deaths attributable to
temperatures.

```{r, eval = TRUE}
## project mortality rates considering historical temperature effect 
proj_rates <- project_mortality_rates(fit_M, fit_F, proj_par)

## project mortality rates considering without temperature effect 
adj_proj_rates <- project_mortality_rates(adj_fit_M, adj_fit_F, adj_proj_par)

# reformat data
format_rates <- function(rates)
{
  rates <- melt(rates)
  names(rates) <- c("age", "year", "sim", "Qxt")
  rates <- rates %>%
    mutate(age_bk = cut(age, age_breaks, age_labels, include.lowest = T))
}

proj_rates <- lapply(proj_rates, format_rates)
names(proj_rates) <- c("m", "f")

adj_proj_rates <- lapply(adj_proj_rates, format_rates)
names(adj_proj_rates) <- c("m", "f")
```

Next, we apply the projected additional mortality related to future temperature.
Estimation errors related to the DLMN model is considered with `r nsim`
bootstrap simulations.

It is worth noting that it is also possible to simultaneously consider
temperature simulations from multiple climate models to capture this source of
uncertainty. However, we do not include this step in this Notebook to simplify
the presentation. Implementing this approach is not difficult and
is been discussed in the paper.

```{r, echo=TRUE, eval = TRUE, fig.width=12, fig.height=8 , message=T, warning=T}

# Change names dates
clim_model <- clim_model %>% dplyr::rename(datedec = date)
# Select the forecasting period
sc <- dplyr::filter(clim_model, years %in% start_year:end_year)

start_time <- Sys.time()
# Run simulations for males and females
res <- lapply(names(adj_proj_rates), function(x)
{
  # Select population
  temp_dem <- adj_proj_rates[[x]] %>%
    filter(year %in% start_year:end_year)
  # Launch forecasting along the rcp scenario with simulations
  # and parallelization
  f_mort <- forecast_mort_dlnm(temp_dem, sc, fit[[x]],
                               q_range = c(q025, q975),
                               nsim = nsim, parallel = parallel,
                               ncpus = ncpus)
  # Add gender indexes
  f_mort <- lapply(f_mort, function(tt){
    if(is.null(tt))
    {
      return(tt)
    } else{
      return(
        tt %>% 
          dplyr::mutate(gender = x) %>%
          relocate(gender, temp_effect, .before =  sim) 
      )
    }
  })
  return(f_mort)
})

# Merge more than two lists with the same element name
res <- Reduce(function(...) Map("rbind", ...), res)

# Save and printtime after operation
end_time <- Sys.time()
time_diff <- end_time - start_time
print(paste("Run time:", as.numeric(time_diff, units = "mins"), "mins"))
```

## Dislaying attributable fraction

In the following figures, we display the projected attributable fraction related
to future temperatures in the selected climate scenario.

To facilitate visual analysis, we calculate an aggregated attributable fraction
to temperatures using the distribution of deaths known at the end of 2019,
as follows
$$
\theta_{t}(D)  = \sum_{g \in \mathcal{G}}
{
\sum_{x \in \mathcal{X}}{\theta_{x,t}^{(g)} (D) \frac{D_{x, 2019}^{(g)}}{D_{2019}}
}},
$$


```{r, eval = TRUE}
# extract the last year 2019, for calculate exposure to risk
expo <- do.call("rbind", lapply(mort_dt$UNI, function(dt){
  dt <- dt$dtx[nrow(dt$dtx),  ]
  return(data.frame(age = 0:105,
                    w = dt,
                    age_bk = c(rep("0-64", 65), rep("65-74", 10),
                               rep("75-84", 10), rep("85+", 21))
                    ))
}))
expo$gender <- c(rep("f", 106), rep("m", 106))
# Agregate expo per age bucket
W <- sum(expo$w)
expo <- expo %>%
  group_by(age_bk, gender) %>%
  summarise(w = sum(w) / W)
```


```{r, fig.width=18, fig.height=12, warning= F, message= F}
# Specify the RCP number
res$tab_excess$traj_clim <- "ALADIN63_CNRM-CM5" 
res$tab_excess$rcp <- "8.5" 
plot_aff <- summary_forecast_temp_effect(res$tab_excess, expo)

# Plot aggregate temperature attributed fraction 
plot_aff[[1]]

p.comb <- plot_aff[[2]] 

# Plot temperature attributed fraction for each age group
p.comb[[1]]
p.comb[[2]]
p.comb[[3]]
p.comb[[4]]
```

## Assessing the loss in life expectancy

Finally, we assess the loss in life expectancy due to future temperatures. 

```{r, fig.width=18, fig.height=12, warning= F, message= F}
res$tab_ex$traj_clim <- "ALADIN63_CNRM-CM5" 
res$tab_ex$rcp <- "8.5" 
plot_ev <- summary_forecast_ev_effect(res$tab_ex)

# Plot LE including temperature effects
plot_ev[[1]]

# Plot lost in LE including temperature effects
plot_ev[[2]]
```


# Conclusion

This Notebook illustrates the implementation of a multi-population mortality
model incorporating the effect of temperature changes on mortality. 
The construction of this framework is simplified by omitting multiple scenarios
and various climate models. Therefore, it does not exactly reproduce the results
presented in the paper.

# References
