---
title: "Results presented in Sections 6.4, 6.5 and in Appendix E and F"
author: "Quentin Guibert"
date: "2025-06-23"
institute: "Université Paris-Dauphine"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
options(encoding = 'UTF-8')
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Folds
folder <- getwd() # If necessary, please set your working directory
fold_data <- paste0(folder, "/data/")
fold_data_raw <- paste0(folder, "/data_raw/")
fold_bib <- paste0(folder, "/functions/")
folderTables <- paste0(folder, "/tables/")
folderFigures <- paste0(folder, "/figures/")
#---------------------------------------------------------------------------------------------------
#### Load functions : ####
# Source functions
for (f in list.files(fold_bib))
  source(paste(fold_bib,f,sep=""), encoding = "UTF-8")

####  Load packages : ####
invisible(lapply(c(
  "lattice",
  "grid",
  "ggplot2",
  "gridExtra",
  "locfit",
  "scales",
  "lubridate",
  "splines",
  "mgcv",
  "dlnm",
  "MASS",
  "mvtnorm",
  "Rfast",
  "parallel",
  "data.table",
  "formattable",
  "RColorBrewer",
  "readxl",
  "kableExtra",
  "dplyr",
  "tidyr",
  "tidyverse",
  "rmarkdown",
  "ggthemes",
  "cowplot",
  "ggridges",
  "viridis",
  "hrbrthemes",
  "colorspace",
  "ggbeeswarm",
  "ggfan",
  "ggpubr"),
  instal.import.package))

#----------------------------------------------------------
#theme graphique
theme_set(theme_bw())
trellis.device(color = FALSE)
```

# Parameters definition

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# General
list_sexe <- list(m = 1,f = 2)
age_breaks <- c(0, 64, 74, 84, Inf)
age_labels <- c("0-64", "65-74", "75-84", "85+")

# training period
annee_deb <- 1980
annee_fin <- 2019

# forecasting period
start_year <- 2020
end_year <- 2100
year_breaks <- c(1980, 1989, 1999, 2009, 2019, 2029, 2039, 2049, 2059, 2069, 2079, 2089, 2100)
year_labels <- c("1980s", "1990s", "2000s", "2010s", "2020s", "2030s", "2040s",
                 "2050s", "2060s", "2070s", "2080s","2090s")

# climate parameters
# list model
info_model <- read_excel("data/info_modeles.xlsx")

# heatwave parameter
min_level <- 17
max_level <- 30
```

## Load data

We calculate weight to summarize attributable fraction (sum across age bucket and sex).

```{r}
load(file =  paste0(fold_data, "/mort_dt.RData"))

# extract the last year 2019, for calculate exposure to risk
expo <- do.call("rbind", lapply(mort_dt$UNI, function(dt){
  dt <- dt$dtx[nrow(dt$dtx),  ]
  return(data.frame(age = 0:105,
                    w = dt,
                    age_bk = c(rep("0-64", 65), rep("65-74", 10), rep("75-84", 10), rep("85+", 21))
  ))
}))
expo$gender <- c(rep("f", 106), rep("m", 106))
# Agregate expo per age bucket
W <- sum(expo$w)
expo <- expo %>%
  group_by(age_bk, gender) %>%
  summarise(w = sum(w) / W)
```

# Results of Section 6.4

## Load results

```{r}
loc_folder <- "adj_proj_rates"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/afftab.RData"))
load(file = paste0(fold_temp_results, "/etab.RData"))
```

## Temperature attributable fraction in Metropolitan France

This code generates Figure 10 of the paper. 

```{r, fig.width=18, fig.height=6}
plot_aff <- summary_forecast_temp_effect(afftab, expo, var = "af_weight_year")

# Display figure
plot_aff[[1]]

# Save figure
sauveFigure(config = 9,
            title = paste0("fig10"),
            path = folderFigures)
plot_aff[[1]]
dev.off()
```

# Results of Section 6.5

## Life expectancy at birth lost in Metropolitan France

This code generates Figure 11 of the paper. 

```{r, fig.width=18, fig.height=12}
plot_ev <- summary_forecast_ev_effect(etab)


# Display figure
plot_ev[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig11"),
            path = folderFigures)
plot_ev[[2]]
dev.off()
```


# Results of Appendix E.1

##  Temperature attributable fraction in Metropolitan France per age group

This code generates Figures 33, 34, 35 and 36 of the paper. 

```{r, fig.width=18, fig.height=12}
p.comb <- plot_aff[[2]]

# Display figure
p.comb[[1]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig33"),
            path = folderFigures)
p.comb[[1]]
dev.off()

# Display figure
p.comb[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig34"),
            path = folderFigures)
p.comb[[2]]
dev.off()

# Display figure
p.comb[[3]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig35"),
            path = folderFigures)
p.comb[[3]]
dev.off()

# Display figure
p.comb[[4]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig36"),
            path = folderFigures)
p.comb[[4]]
dev.off()
```


# Results of Appendix E.2

##  Temperature attributable fraction in Metropolitan France per city

This code generates Figures 37 of the paper. 

```{r, fig.width=18, fig.height=6}

# Load results
loc_folder <- "adj_proj_rates_Brest"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/afftab.RData"))

plot_aff <- summary_forecast_temp_effect(afftab, expo, var = "af_weight_year")

# Display figure
plot_aff[[1]]

# Save figure
sauveFigure(config = 9,
            title = paste0("fig37"),
            path = folderFigures)
plot_aff[[1]]
dev.off()
```

This code generates Figures 38 of the paper. 

```{r, fig.width=18, fig.height=6}

# Load results
loc_folder <- "adj_proj_rates_Marseille"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/afftab.RData"))

plot_aff <- summary_forecast_temp_effect(afftab, expo, var = "af_weight_year")

# Display figure
plot_aff[[1]]

# Save figure
sauveFigure(config = 9,
            title = paste0("fig38"),
            path = folderFigures)
plot_aff[[1]]
dev.off()
```

This code generates Figures 39 of the paper. 

```{r, fig.width=18, fig.height=6}

# Load results
loc_folder <- "adj_proj_rates_Paris"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/afftab.RData"))

plot_aff <- summary_forecast_temp_effect(afftab, expo, var = "af_weight_year")

# Display figure
plot_aff[[1]]

# Save figure
sauveFigure(config = 9,
            title = paste0("fig39"),
            path = folderFigures)
plot_aff[[1]]
dev.off()
```

This code generates Figures 40 of the paper. 

```{r, fig.width=18, fig.height=6}

# Load results
loc_folder <- "adj_proj_rates_Perpignan"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/afftab.RData"))

plot_aff <- summary_forecast_temp_effect(afftab, expo, var = "af_weight_year")

# Display figure
plot_aff[[1]]

# Save figure
sauveFigure(config = 9,
            title = paste0("fig40"),
            path = folderFigures)
plot_aff[[1]]
dev.off()
```

This code generates Figures 41 of the paper. 

```{r, fig.width=18, fig.height=6}

# Load results
loc_folder <- "adj_proj_rates_Strasbourg"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/afftab.RData"))

plot_aff <- summary_forecast_temp_effect(afftab, expo, var = "af_weight_year")

# Display figure
plot_aff[[1]]

# Save figure
sauveFigure(config = 9,
            title = paste0("fig41"),
            path = folderFigures)
plot_aff[[1]]
dev.off()
```

# Results of Appendix F

## Life expectancy at birth lost per city

This code generates Figure 42 of the paper. 

```{r, fig.width=18, fig.height=12}
# Load results
loc_folder <- "adj_proj_rates_Brest"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/etab.RData"))

plot_ev <- summary_forecast_ev_effect(etab)

# Display figure
plot_ev[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig42"),
            path = folderFigures)
plot_ev[[2]]
dev.off()
```


This code generates Figure 43 of the paper. 

```{r, fig.width=18, fig.height=12}
# Load results
loc_folder <- "adj_proj_rates_Marseille"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/etab.RData"))

plot_ev <- summary_forecast_ev_effect(etab)

# Display figure
plot_ev[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig43"),
            path = folderFigures)
plot_ev[[2]]
dev.off()
```

This code generates Figure 44 of the paper. 

```{r, fig.width=18, fig.height=12}
# Load results
loc_folder <- "adj_proj_rates_Paris"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/etab.RData"))

plot_ev <- summary_forecast_ev_effect(etab)

# Display figure
plot_ev[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig44"),
            path = folderFigures)
plot_ev[[2]]
dev.off()
```

This code generates Figure 45 of the paper. 

```{r, fig.width=18, fig.height=12}
# Load results
loc_folder <- "adj_proj_rates_Perpignan"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/etab.RData"))

plot_ev <- summary_forecast_ev_effect(etab)

# Display figure
plot_ev[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig45"),
            path = folderFigures)
plot_ev[[2]]
dev.off()
```

This code generates Figure 46 of the paper. 

```{r, fig.width=18, fig.height=12}
# Load results
loc_folder <- "adj_proj_rates_Strasbourg"
fold_temp_results <- paste0(folder, "/data/", loc_folder)
load(file = paste0(fold_temp_results, "/etab.RData"))

plot_ev <- summary_forecast_ev_effect(etab)

# Display figure
plot_ev[[2]]

# Save figure
sauveFigure(config = 5,
            title = paste0("fig46"),
            path = folderFigures)
plot_ev[[2]]
dev.off()
```

# Session Information

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```
