---
title: "Results presented in Section 6.1 and in Appendix B"
author: "Quentin Guibert"
date: "2025-06-23"
institute: "Université Paris-Dauphine"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
options(encoding = 'UTF-8')
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Folds
folder <- getwd() # If necessary, please set your working directory
fold_data <- paste0(folder, "/data/")
fold_data_raw <- paste0(folder, "/data_raw/DRIAS/")
fold_bib <- paste0(folder, "/functions/")
folderTables <- paste0(folder, "/tables/")
folderFigures <- paste0(folder, "/figures/")
#---------------------------------------------------------------------------------------------------
#### Load functions : ####
# Source functions
for (f in list.files(fold_bib))
  source(paste(fold_bib,f,sep=""), encoding = "UTF-8")

####  Load packages : ####
invisible(lapply(c(
  "lattice",
  "grid",
  "ggplot2",
  "gridExtra",
  "locfit",
  "scales",
  "lubridate",
  "splines",
  "mgcv",
  "data.table",
  "formattable",
  "RColorBrewer",
  "readxl",
  "kableExtra",
  "gtsummary",
  "dplyr",
  "tidyr",
  "tidyverse",
  "rmarkdown",
  "ggthemes",
  "ggforce",
  "cowplot",
  "ggridges",
  "viridis",
  "hrbrthemes",
  "colorspace",
  "ggbeeswarm",
  "ggfan"),
  instal.import.package))

#----------------------------------------------------------
#theme graphique
theme_set(theme_bw())
trellis.device(color = FALSE)
```

# Parameters definition

```{r, echo=TRUE, message=FALSE, warning=FALSE}
### Parameters
#---------------------------------------------------------------------------------------------------
list_sexe <- list(m = 1,f = 2)
age_breaks <- c(0, 64, 74, 84, Inf)
age_labels <- c("0-64", "65-74", "75-84", "85+")

# training period
annee_deb <- 1980
annee_fin <- 2019
outlier <- 999 # exclude no atypical year

# other parameters
AGE_MAX <- 108
```

# Load data series 

We import daily data series for death counts and temperatures.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
base_nb_deces <- lapply(list_sexe, function(s){
  import_data_death(s, annee_deb, annee_fin, outlier)
})
base_nb_deces <- do.call("rbind", base_nb_deces)
```

# Summary statistics

This code generates Figure 1 of the paper. 

## Distribution of average daily temperatures for each month of the year
```{r, echo=TRUE, message=FALSE, warning=FALSE}
# ajust layout
des.1 <- base_nb_deces %>%
  mutate(month = as.factor(month)) %>%
  filter(age_bk == "0-49") # only one age group is necessary

# add month format
des.1$month <- fct_recode(des.1$month, "Jan." = "1",
                          "Feb." ="2",
                          "Mar." ="3",
                          "Apr." = "4",
                          "May" = "5",
                          "June" = "6",
                          "July" = "7",
                          "Aug." = "8",
                          "Sept." = "9",
                          "Oct." = "10",
                          "Nov." = "11",
                          "Dec." = "12")

# plot geom_density_ridges()
g <- ggplot(
  des.1, 
  aes(x = TAVG, y = month, fill = stat(x))) +
  geom_density_ridges_gradient(scale = 3, size = 0.3, rel_min_height = 0.01) +
  scale_fill_continuous_diverging(palette = "Blue-Yellow 3") +
  theme_bw() +
  labs(x = "Avg. Temperature [C]",
       y = "Months",
       fill = "Temp. [C]")

# Display figure
g

# save figure
sauveFigure(config = 7, 
            title = paste0("fig1"),
            path = folderFigures)
g
dev.off()
```


## Representation of the daily death counts according to the average daily temperature in France for women and men

This code generates Figure 2 of the paper. 
```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.height = 6, fig.width = 12}

# ajust layout
des.2 <- base_nb_deces %>%
  mutate(sexe = as.factor(sexe)) %>%
  mutate(month = as.factor(month)) %>%
  select(DateDec, Nombre_de_deces, TAVG, sexe) %>%
  group_by(DateDec, sexe) %>%
  summarise(Nombre_de_deces = sum(Nombre_de_deces),
            TAVG = mean(TAVG))

des.2$sexe <- fct_recode(des.2$sexe, "Males" = "1",
                         "Females" ="2")

#  plot
g <- ggplot(
  des.2, 
  aes(x = TAVG, y = Nombre_de_deces)) +
  geom_point(size = 0.5) +
  geom_smooth() +
  facet_wrap(~ sexe) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA)) +
  labs(x = "Avg. Temperature [C]",
       y = "Daily deaths")

# Display figure
g

# save figure
sauveFigure(config = 8, 
            title = paste0("fig2"),
            path = folderFigures)
g
dev.off()
```


## Probability density of the number of deaths by month of the year

This code generates Figure 3 of the paper. 

```{r, echo=TRUE, message=FALSE, warning=FALSE,fig.height = 6, fig.width = 12}
# ajust layout
des.3 <- des.2
des.3$month <- as.factor(month(des.2$DateDec))

# add month format
des.3$month <- fct_recode(des.3$month, "Jan." = "1",
                          "Feb." ="2",
                          "Mar." ="3",
                          "Apr." = "4",
                          "May" = "5",
                          "June" = "6",
                          "July" = "7",
                          "Aug." = "8",
                          "Sept." = "9",
                          "Oct." = "10",
                          "Nov." = "11",
                          "Dec." = "12")

#  geom_density_ridges()
g <- ggplot(
  des.3, 
  aes(x = month, y = Nombre_de_deces, fill = month)) +
  geom_violin(alpha = 0.5) +
  # geom_point(position = position_jitter(seed = 1, width = 0.2), size = 0.1) +
  stat_summary(fun.data=mean_sdl, mult=1, 
               geom="pointrange", color="black") +
  facet_wrap(~ sexe) +
  theme_bw() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA)) +
  labs(x = "Months",
       y = "Daily deaths")

# Display figure
g

# save figure
sauveFigure(config = 8, 
            title = paste0("fig3"),
            path = folderFigures)
g
dev.off()
```

# Climate scenario dataset

## Load data series
We import daily data series for climate trajectories. The models selected here
are those with both RCP 2.6, 4.5 and RCP 8.5 projections.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

# import list model
info_model <- read_excel("data/info_modeles.xlsx")

# period parameters
start_year <- 2020
end_year <- 2100
year_breaks <- c(2000, 2009, 2019, 2029, 2039, 2049, 2059, 2069, 2079, 2089, 2100)
year_labels <- c("2000s", "2010s", "2020s", "2030s", "2040s",
                 "2050s", "2060s", "2070s", "2080s","2090s")
# heatwave parameter
min_level <- 17
max_level <- 30

# import data
model <- lapply(1:nrow(info_model), function(i)
{
  print(paste0(info_model$Id[i],"- RCP", info_model$RCP[i]))
  treat_rcp(info_model$Id[i], info_model$RCP[i], info_model,
            min_level, max_level, start_year, end_year, 
            new_calcul= FALSE, city = FALSE, path = fold_data, path_raw = fold_data_raw)
})
model <- do.call("rbind", model)
```

## Projection of temperatures and heatwaves by RCP scenario in Metropolitan France over the period 2020-2100

This code generates Figure 4 of the paper. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
figures <- summary_climate(model)
```

```{r fig_proj_temp_rcp, echo=TRUE, message=FALSE, warning=FALSE,fig.height = 12, fig.width = 12}
# Display figure
figures$fig_agg

# save figure
sauveFigure(config = 7, 
            title = paste0("fig4"),
            path = folderFigures)
figures$fig_agg
dev.off()
```

## Additional informations in Section 6.1

The figures presented in the last paragraph of Section 6.1 can be reproduced by running the code below. 

```{r}
# Some statistics
df <- model %>%
  filter(YEARS >= start_year, YEARS < end_year) %>%
  group_by(model, rcp, YEARS, Num_vague) %>%
  summarise(Duration = sum(Ind_canicule),
            Severity = sum(Severite_canicule),
            Intensity = max(Severite_canicule)) %>%
  group_by(model, rcp) %>%
  summarise(nv_vag = sum(Num_vague),
            max_dur = max(Duration),
            max_dur_name = YEARS[which.max(Duration)],
            max_sev = max(Severity),
            max_sev_name = YEARS[which.max(Severity)],
            max_int = max(Intensity),
            max_int_name = YEARS[which.max(Intensity)]
  )


df <- df %>%
  rename(
    Model = model,
    RCP = rcp ,
    Heatwave_Counts = nv_vag,
    Max_Duration = max_dur,
    Year_of_Max_Duration = max_dur_name,
    Max_Severity = max_sev,
    Year_of_Max_Severity = max_sev_name,
    Max_Intensity = max_int,
    Year_of_Max_Intensity = max_int_name 
  )
  
# Filter on scenario RCP 2.6
df %>%
  filter(RCP == "2.6") %>%  
  kableExtra::kable(booktabs=TRUE) %>%
  kable_classic_2(full_width = F)

# Filter on scenario  RCP 4.5
df %>%
  filter(RCP == "4.5") %>%
  kableExtra::kable(booktabs=TRUE) %>%
  kable_classic_2(full_width = F)

# Filter on scenario RCP 8.5
df %>%
  filter(RCP == "8.5") %>%
  kableExtra::kable(booktabs=TRUE) %>%
  kable_classic_2(full_width = F)
```

# Results of Appendix B

## Load and prepare daily data 

```{r, echo=TRUE}
base_nb_deces <- lapply(list_sexe, function(s){
  import_data_death(s, annee_deb, annee_fin, outlier)
})

base_nb_deces <- lapply(base_nb_deces, function(data){
  data$age_bk <- ifelse(
    data$age_bk == "0-49", "0-64", ifelse(
      data$age_bk == "50-54", "0-64", ifelse(
        data$age_bk == "55-59", "0-64", ifelse(
          data$age_bk == "60-64", "0-64", ifelse(
            data$age_bk == "65-69", "65-74", ifelse(
              data$age_bk == "70-74", "65-74", ifelse(
                data$age_bk == "75-79", "75-84", ifelse(
                  data$age_bk == "80-84", "75-84", ifelse(
                    data$age_bk == "85-89", "85+", ifelse(
                      data$age_bk == "90 et +", "85+", NA
                    ))))))))))
  
  data <- data %>%
    group_by(DateDec, age_bk) %>%
    summarise(Nombre_de_deces = sum(Nombre_de_deces),
              YEARS = min(YEARS),
              TAVG = min(TAVG),
              Ind_canicule = min(Ind_canicule),
              time = min(time),
              dow = min(dow),
              month = min(month)
    )
  return(as.data.table(data))
})

# Aggregate number of deaths by cluster
## Daily data
tab_agg_daily_death <- do.call("rbind", lapply(c("f", "m"), function(ss)
{
 df <- base_nb_deces[[ss]] %>%
   dplyr::select(DateDec, age_bk, Nombre_de_deces, YEARS)%>%
   mutate(sex = ss) %>%
   group_by(YEARS, age_bk, sex) %>%
   summarise(deaths = sum(Nombre_de_deces)) %>% 
   ungroup()
}))
```

## Load and prepare annual data 

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F}

# Get death and exposure data from HMD
load(file =  paste0(fold_data, "/dtx_fr.RData"))

## Annual mortality
dtx_fr$age_bk <- ifelse(dtx_fr$Age <65 , "0-64",
                        ifelse(dtx_fr$Age <75 ,"65-74",
                               ifelse(dtx_fr$Age <85 ,"75-84","85+")
                        ))
tab_agg_annual_death <- dtx_fr %>%
  dplyr::select(Year, Age, age_bk, Female, Male) %>%
  pivot_longer(cols = Female:Male,
               names_to = "sex", values_to = "deaths") %>%
  mutate(sex = factor(sex)) %>%
  group_by(Year, age_bk, sex) %>%
  summarise(deaths = sum(deaths)) %>% 
   ungroup()

tab_agg_annual_death$sex <- fct_recode(tab_agg_annual_death$sex,
                                       f = "Female", m = "Male")
```

## Comparison of annual death counts from daily data and annual data

This code generates Figure 12 of the paper. 

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F}  
# Merge data
tab_compar_deaths <- tab_agg_daily_death %>%
  left_join(tab_agg_annual_death, by = c("YEARS"= "Year",
                                         "age_bk"= "age_bk",
                                         "sex" = "sex")) %>%
  mutate(age_bk = factor(age_bk), gap = (deaths.x - deaths.y) / deaths.x)%>%
  pivot_longer(cols = deaths.x:deaths.y,
               names_to = "source", values_to = "deaths")

tab_compar_deaths$source <- fct_recode(tab_compar_deaths$source,
                                       daily = "deaths.x", yearly = "deaths.y")  


# Compar death counts by age bucket and sex
p_compar_deaths <- ggplot(tab_compar_deaths, aes(x = YEARS, y = deaths)) +
  geom_line(aes(color = source)) +
  facet_grid(sex ~ age_bk, labeller = labeller(sex = c("f" = "Female", "m" = "Male")))+ 
  labs(x = "Year",
       y = "Number of deaths") +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = c(0.05, 0.90),
    legend.background = element_rect(fill = "white", color = "black"),
    legend.title = element_blank()
  )

# Display figure
p_compar_deaths

# save figure
sauveFigure(config = 11, 
            title = paste0("fig12"),
            path = folderFigures)
p_compar_deaths
dev.off()
```


# Session Information

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```
