---
title: "Results presented in Section 6.2 and in Appendix C.1 and C.2"
author: "Quentin Guibert"
date: "2025-06-23"
institute: "Université Paris-Dauphine"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
options(encoding = 'UTF-8')
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Folds
folder <- getwd() # If necessary, please set your working directory
fold_data <- paste0(folder, "/data/")
fold_data_raw <- paste0(folder, "/data_raw/")
fold_bib <- paste0(folder, "/functions/")
folderTables <- paste0(folder, "/tables/")
folderFigures <- paste0(folder, "/figures/")
#---------------------------------------------------------------------------------------------------
#### Load functions : ####
# Source functions
for (f in list.files(fold_bib))
  source(paste(fold_bib,f,sep=""), encoding = "UTF-8")

####  Load packages : ####
invisible(lapply(c(
  "lattice",
  "grid",
  "ggplot2",
  "gridExtra",
  "locfit",
  "scales",
  "lubridate",
  "splines",
  "mgcv",
  "dlnm",
  "MASS",
  "mvtnorm",
  "Rfast",
  "parallel",
  "data.table",
  "formattable",
  "RColorBrewer",
  "readxl",
  "kableExtra",
  "dplyr",
  "tidyr",
  "tidyverse",
  "rmarkdown",
  "ggthemes",
  "cowplot",
  "ggridges",
  "viridis",
  "hrbrthemes",
  "colorspace",
  "ggbeeswarm",
  "ggfan"),
  instal.import.package))

#----------------------------------------------------------
#theme graphique
theme_set(theme_bw())
trellis.device(color = FALSE)
```

# Parameters definition

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Main parameters
list_sexe <- list(m = 1,f = 2)
age_breaks <- c(0, 64, 74, 84, Inf)
age_labels <- c("0-64", "65-74", "75-84", "85+")

# training period
annee_deb <- 1980
annee_fin <- 2019
annee_in_sample <- 1980:2015
annee_out_sample <- 2016:2019
outlier <- 999 # exclude no atypical year

# forecasting period
start_year <- 2020
end_year <- 2070
year_breaks <- c(1980, 1989, 1999, 2009, 2019, 2029, 2039, 2049, 2059, 2069, 2079, 2089, 2100)
year_labels <- c("1980s", "1990s", "2000s", "2010s", "2020s", "2030s", "2040s",
                 "2050s", "2060s", "2070s", "2080s","2090s")

config_temp <-c(
  "all_effect", "hot_effect", "cold_effect", "moderate_hot_effect",
  "moderate_cold_effect","extr_hot_effect", "extr_cold_effect", "canicule_effect"
)

# simulation
nsim <- 100 # number of simulations
parallel <- "no" # parallel option
ncpus <- 4 # number of cores

# other parameters
AGE_MAX <- 108
```

# Load data series 

We import daily data series for death counts and temperatures. In the following lines,
we prepare our data before fiting the DLNM model. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
base_nb_deces <- lapply(list_sexe, function(s){
  import_data_death(s, annee_deb, annee_fin, outlier)
})

base_nb_deces <- lapply(base_nb_deces, function(data){
  data$age_bk <- ifelse(
    data$age_bk == "0-49", "0-64", ifelse(
      data$age_bk == "50-54", "0-64", ifelse(
        data$age_bk == "55-59", "0-64", ifelse(
          data$age_bk == "60-64", "0-64", ifelse(
            data$age_bk == "65-69", "65-74", ifelse(
              data$age_bk == "70-74", "65-74", ifelse(
                data$age_bk == "75-79", "75-84", ifelse(
                  data$age_bk == "80-84", "75-84", ifelse(
                    data$age_bk == "85-89", "85+", ifelse(
                      data$age_bk == "90 et +", "85+", NA
                    ))))))))))
  
  data <- data %>%
    group_by(DateDec, age_bk) %>%
    summarise(Nombre_de_deces = sum(Nombre_de_deces),
              YEARS = min(YEARS),
              TAVG = min(TAVG),
              Ind_canicule = min(Ind_canicule),
              time = min(time),
              dow = min(dow),
              month = min(month)
    )
  return(as.data.table(data))
})

```

# DLNM model for the whole year with LAG = 21 days

We fit DLNM models for each gender and each age bucket. These models are calibrated
over all the years. 

## Parameters definition
```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F}
# DLNM Parameters
param_dlnm_whole <- list(
  # main model, cubic natural spline with three internal knots in
  #   the 10th, 75th, 90th percentiles of the temperature distribution
  varfun = "bs",
  vardegree = 2,
  varper  = c(10,75,90),
  ## specification of the lag function
  # Definition of the maximum lag, that is, 21 days
  lag  = 21,
  lagnk  = 3,
  ## degree of freedom for seasonality
  dfseas  = 8,
  ## degree of freedom for trend
  dftrend = NULL
)
```

## Model estimation

In the following lines, we fit the DLNM model.

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F, comment= F}
fit <- lapply(base_nb_deces, function(x){
  fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)
})
```

Then, we perform in-sample analysis.

```{r echo=TRUE, fig.width=12, fig.height=8}
# Produce the RR plots
tempplot_all_sex <- summary_dlnm_all_sex(fit)

# in-sample performance of the fitted model for the whole year.
perf_mort <- lapply(names(base_nb_deces), function(x){
  perf_dnlm(fit[[x]])
})
names(perf_mort) <- names(base_nb_deces)
```

## Calculating attribable fraction

```{r, echo=TRUE, eval = TRUE, fig.width=12, fig.height=8}
# Quantiles for extreme hot and cold
q025 <- quantile(base_nb_deces$m$TAVG, 0.025)
q975 <- quantile(base_nb_deces$m$TAVG, 0.975)

# Run Excess of mortality
xs_mort <- lapply(names(base_nb_deces), function(x){
  res <- excess_mort_dlnm(base_nb_deces[[x]],
                          fit[[x]],
                          q_range = c(q025, q975),
                          nsim = nsim, 
                          parallel = parallel, 
                          ncpus = ncpus)
  return(res)
})
names(xs_mort) <- names(base_nb_deces)

# Save the results
save(xs_mort, file =  paste0(fold_data, "xs_mort_lag_21"))
```

# Results of Section 6.2

## Cumulative relative risk of mortality over a 21-day period in Metropolitan France

This code generates Figure 5 of the paper. 

```{r dlnm_rr_curve, eval = TRUE, echo=TRUE,fig.height = 12, fig.width = 12}
# Display figure
tempplot_all_sex

# save figure
sauveFigure(config = 7,
            title = paste0("fig5"),
            path = folderFigures)
tempplot_all_sex
dev.off()
```

## Temperature attributable fraction in Metropolitan France, 

This code generates Figure 6 of the paper. 

```{r temp_decomp, echo=TRUE, fig.width=12, fig.height=6}

# Convert to a data frame
xs_data <- lapply(names(xs_mort), function(i){
  xs_mort[[i]]$excess_mort %>%
    dplyr::mutate(gender = i) %>%
    rename(year = YEARS)
})
xs_data <- do.call("rbind", xs_data)


# Generate plot
sub_palette <- c("black", "blue", "red", "lightblue", "orange")
p <- xs_data %>%
  dplyr::filter(temp_effect %in% c("all_effect",
                                   "moderate_hot_effect",
                                   "moderate_cold_effect",
                                   "extr_hot_effect","extr_cold_effect")) %>%
  mutate(temp_effect = fct_recode(temp_effect,
                                  "All effects" = "all_effect",
                                  "Extreme cold" = "extr_cold_effect",
                                  "Extreme hot" = "extr_hot_effect",
                                  "Moderate cold" = "moderate_cold_effect",
                                  "Moderate hot" = "moderate_hot_effect")) %>%
  mutate(gender = fct_recode(gender, "Females" = "f" ,
                             "Males" = "m")) %>%
  group_by(year, sim, temp_effect, gender) %>%
  summarise(Dxt = sum(Dxt),
            an_sum_all_ages = sum(an_sum),
            cases_sum_all_ages = sum(cases_sum)) %>%
  mutate(xs_rate = an_sum_all_ages / cases_sum_all_ages * 100) %>%
  group_by(year, temp_effect, gender) %>%
  summarise(xsmean = mean(xs_rate),
            xsmax = quantile(xs_rate, 0.95),
            xsmin = quantile(xs_rate, 0.05)) %>%
  ggplot(aes(x = year, y = xsmean)) +
  geom_line(aes(colour = factor(temp_effect))) +
  geom_ribbon(aes(ymax = xsmax, ymin = xsmin, fill = factor(temp_effect)), alpha = 0.3) +
  facet_wrap(~ gender, scales = "free_x", ncol = 2) + 
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = "white", color = "black"),
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))+
  labs(x = "Years", y = "Excess mortality (%)", fill = "Temperature effect", 
       colour = "Temperature effect") +
  scale_colour_manual(values = sub_palette) +
  scale_fill_manual(values = sub_palette)

# Display figure
p

# save figure
sauveFigure(config = 8, 
            title = paste0("fig6"),
            path = folderFigures)
p
dev.off()
```


# Results of Appendix C.1

In this section, we present the predicted all-cause death counts obtained from the DLNM model,
and visually assess its ability to reproduce the seasonality of daily deaths
across age groups and genders.

## Comparison of observed daily deaths and predicted deaths for women

This code generates Figure 13 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=12 }
# Display figure
perf_mort$f$desc_1

# save figure
sauveFigure(config = 7,
            title = paste0("fig13"),
            path = folderFigures)
perf_mort$f$desc_1
dev.off()
```

## Comparison of observed daily deaths and predicted deaths for men

This code generates Figure 14 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$m$desc_1

# save figure
sauveFigure(config = 7,
            title = paste0("fig14"),
            path = folderFigures)
perf_mort$m$desc_1
dev.off()
```

## Decomposition of daily death counts for women predicted by the DLNM model

This code generates Figure 15 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$f$desc_2

# save figure
sauveFigure(config = 7,
            title = paste0("fig15"),
            path = folderFigures)
perf_mort$f$desc_2
dev.off()
```


## Decomposition of daily death counts for men predicted by the DLNM model

This code generates Figure 16 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$m$desc_2

# save figure
sauveFigure(config = 7,
            title = paste0("fig16"),
            path = folderFigures)
perf_mort$m$desc_2
dev.off()
```

## Evolution of daily average temperatures

This code generates Figure 17 of the paper. 

```{r temperature, eval = TRUE, echo=TRUE,fig.height = 12, fig.width = 12}
p_temperature  <- ggplot(fit$m[[1]]$model$data) +
  geom_line(aes(x = DateDec, y = TAVG), size = 0.5, col = "black") +
  xlab("Year") + ylab("Avg. Temperature [C]") +
  theme_bw() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))

# Display figure
p_temperature 

# save figure
sauveFigure(config = 1,
            title = paste0("fig17"),
            path = folderFigures)
p_temperature
dev.off()
```

# Results of Appendix C.2

In this section, we conduct a statistical analysis to evaluate the goodness
of fit of the DLNM model.

## Monthly distribution of observed and predicted numbers of deaths based for women

This code generates Figure 18 of the paper. 

```{r dlnm_obs_pred_f, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$f$perf

# save figure
sauveFigure(config = 7,
            title = paste0("fig18"),
            path = folderFigures)
perf_mort$f$perf
dev.off()
```

## Monthly distribution of observed and predicted numbers of deaths based for men

This code generates Figure 19 of the paper. 

```{r dlnm_obs_pred_m, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$m$perf

# save figure
sauveFigure(config = 7,
            title = paste0("fig19"),
            path = folderFigures)
perf_mort$m$perf
dev.off()
```

## Representation of deviance residuals for DLNM models for women

This code generates Figure 20 of the paper. 

```{r dlnm_resid_f, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$f$residual

# save figure
sauveFigure(config = 10,
            title = paste0("fig20"),
            path = folderFigures)
perf_mort$f$residual
dev.off()
```


## Representation of deviance residuals for DLNM models for men

This code generates Figure 21 of the paper. 

```{r dlnm_resid_m, echo=TRUE, fig.width=12, fig.height=12}
# Display figure
perf_mort$m$residual

# save figure
sauveFigure(config = 10,
            title = paste0("fig21"),
            path = folderFigures)
perf_mort$m$residual
dev.off()
```

# Session Information

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```
