---
title: "Results presented Appendix C.3"
author: "Quentin Guibert"
date: "2025-06-23"
institute: "Université Paris-Dauphine"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
options(encoding = 'UTF-8')
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Folds
folder <- getwd() # If necessary, please set your working directory
fold_data <- paste0(folder, "/data/")
fold_data_raw <- paste0(folder, "/data_raw/")
fold_bib <- paste0(folder, "/functions/")
folderTables <- paste0(folder, "/tables/")
folderFigures <- paste0(folder, "/figures/")
#---------------------------------------------------------------------------------------------------
#### Load functions : ####
# Source functions
for (f in list.files(fold_bib))
  source(paste(fold_bib,f,sep=""), encoding = "UTF-8")

####  Load packages : ####
invisible(lapply(c(
  "lattice",
  "grid",
  "ggplot2",
  "gridExtra",
  "locfit",
  "scales",
  "lubridate",
  "splines",
  "mgcv",
  "dlnm",
  "MASS",
  "mvtnorm",
  "Rfast",
  "parallel",
  "data.table",
  "formattable",
  "RColorBrewer",
  "readxl",
  "kableExtra",
  "dplyr",
  "tidyr",
  "tidyverse",
  "rmarkdown",
  "ggthemes",
  "cowplot",
  "ggridges",
  "viridis",
  "hrbrthemes",
  "colorspace",
  "ggbeeswarm",
  "ggfan"),
  instal.import.package))

#----------------------------------------------------------
#theme graphique
theme_set(theme_bw())
trellis.device(color = FALSE)
```

# Parameters definition

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Main parameters
list_sexe <- list(m = 1,f = 2)
age_breaks <- c(0, 64, 74, 84, Inf)
age_labels <- c("0-64", "65-74", "75-84", "85+")

# training period
annee_deb <- 1980
annee_fin <- 2019
annee_in_sample <- 1980:2015
annee_out_sample <- 2016:2019
outlier <- 999 # exclude no atypical year

# forecasting period
start_year <- 2020
end_year <- 2070
year_breaks <- c(1980, 1989, 1999, 2009, 2019, 2029, 2039, 2049, 2059, 2069, 2079, 2089, 2100)
year_labels <- c("1980s", "1990s", "2000s", "2010s", "2020s", "2030s", "2040s",
                 "2050s", "2060s", "2070s", "2080s","2090s")

config_temp <-c(
  "all_effect", "hot_effect", "cold_effect", "moderate_hot_effect",
  "moderate_cold_effect","extr_hot_effect", "extr_cold_effect", "canicule_effect"
)

# simulation
nsim <- 10 # number of simulations
parallel <- "no" # parallel option
ncpus <- 4 # number of cores

# other parameters
AGE_MAX <- 108
```

# Load data series 

We import daily data series for death counts and temperatures. In the following lines,
we prepare our data before fiting the DLNM model. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
base_nb_deces <- lapply(list_sexe, function(s){
  import_data_death(s, annee_deb, annee_fin, outlier)
})

base_nb_deces <- lapply(base_nb_deces, function(data){
  data$age_bk <- ifelse(
    data$age_bk == "0-49", "0-64", ifelse(
      data$age_bk == "50-54", "0-64", ifelse(
        data$age_bk == "55-59", "0-64", ifelse(
          data$age_bk == "60-64", "0-64", ifelse(
            data$age_bk == "65-69", "65-74", ifelse(
              data$age_bk == "70-74", "65-74", ifelse(
                data$age_bk == "75-79", "75-84", ifelse(
                  data$age_bk == "80-84", "75-84", ifelse(
                    data$age_bk == "85-89", "85+", ifelse(
                      data$age_bk == "90 et +", "85+", NA
                    ))))))))))
  
  data <- data %>%
    group_by(DateDec, age_bk) %>%
    summarise(Nombre_de_deces = sum(Nombre_de_deces),
              YEARS = min(YEARS),
              TAVG = min(TAVG),
              Ind_canicule = min(Ind_canicule),
              time = min(time),
              dow = min(dow),
              month = min(month)
    )
  return(as.data.table(data))
})
```

# Results of Appendix C.3

## Degrees of freedom and knots hyperparameters

We examine the sensitivity of the DLNM model to the number of internal knots for the natural cubic
B-spline in the lag-response curve, as well as to the degrees of freedom which controls the
residual seasonality.

By running the script `7-Main_Generate_Sensi.R`, we performs different sensibility to hyperparameters.

```{r, echo=TRUE, fig.width=12, fig.height=8}
load(file =  paste0(fold_data, "/sensi_results.RData"))

sensi_results$quant_knot_text <- unlist(paste(sensi_results$quant_knot))
# Remove NA
sensi_results <- sensi_results %>%
  na.omit()

# Order raws
sensi_results <- sensi_results %>%
  group_by(sex, age_bk) %>%
  arrange(qbic, qaic) %>%
  dplyr::select(quant_knot_text, nk, seas, sex, age_bk, qaic, qbic)

# Aggregate by parameters
sensi_agg <- sensi_results %>%
  mutate(quant_knot_text = factor(quant_knot_text)) %>%
  group_by(quant_knot_text, nk, seas) %>%
  summarise(qaic_sum = sum(qaic),
            qbic_sum = sum(qbic)) %>%
  arrange(qbic_sum, qaic_sum)

# Aggregate by parameters and sex
sensi_agg_f <- sensi_results %>%
  dplyr::filter(sex == "f") %>%
  mutate(quant_knot_text = factor(quant_knot_text)) %>%
  group_by(quant_knot_text, nk, seas) %>%
  summarise(qaic_sum = sum(qaic),
            qbic_sum = sum(qbic)) %>%
  arrange(qbic_sum, qaic_sum)


sensi_agg_m <- sensi_results %>%
  dplyr::filter(sex == "m") %>%
  mutate(quant_knot_text = factor(quant_knot_text)) %>%
  group_by(quant_knot_text, nk, seas) %>%
  summarise(qaic_sum = sum(qaic),
            qbic_sum = sum(qbic)) %>%
  arrange(qbic_sum, qaic_sum)

# Finally, I remove  quant_knot_text with 4 points
sensi_agg <- sensi_agg %>%
  filter(quant_knot_text == "c(10, 75, 90)")
sensi_agg_f <- sensi_agg_f %>%
  filter(quant_knot_text == "c(10, 75, 90)")
sensi_agg_m <- sensi_agg_m %>%
  filter(quant_knot_text == "c(10, 75, 90)")
```

## Comparison of QAIC and QBIC statistics for the 10 best DLNM models

We now compare results and select the best model. This code generates Table 5 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8}

# Create tab with the 10 best models
tab_sensi <- sensi_agg[1:10, ] %>%
  ungroup() %>%
  mutate(model = paste0("Model ", 1:10),
         quant_knot_text = NULL)  %>%
  relocate(model) 

# Print Table
tab_sensi %>%
  rename(
    Model = model,
    Number_knot = nk,
    Seanonality_degree = seas,
    QAIC = qaic_sum,
    QBIC = qbic_sum
  ) %>%
  kableExtra::kable(booktabs=TRUE) %>%
  kable_classic_2(full_width = F)

# Save table in .tex
tab_sensi  %>%
  kbl(booktabs = T, format = 'latex', 
      align='c',
      col.names = c("Model","Number of knots (lag-response)", "Seanonality degree", "QAIC", "QBIC"),
      digits=2, 
      escape =  FALSE) %>%
  kable_classic_2(full_width = F) %>% 
  kable_styling(latex_options = "hold_position",  font_size = 8)%>%
  readr::write_lines(file = paste0(folderTables, "tab5.tex"))
```


## Cumulative relative risk of mortality over a 21-day period for women

We refit the 10 best DLNM models and compare results using the cumulative
RR of mortality. 

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F, comment= F}

## Compar RR curve
rr_curve_sensi <- lapply(1:10, function(i){
  
  #Parameter
  param_dlnm_whole <- list(
    # main model, cubic natural spline with three internal knots in
    #   the 10th, 75th, 90th percentiles of the temperature distribution
    varfun = "bs",
    vardegree = 2,
    varper  = c(10, 75, 90),
    ## specification of the lag function
    # Definition of the maximum lag, that is, 21 days
    lag  = 21,
    lagnk  = sensi_agg$nk[i],
    ## degree of freedom for seasonality
    dfseas  = sensi_agg$seas[i],
    ## degree of freedom for trend
    dftrend = NULL
  )
  
  fit <- lapply(base_nb_deces, function(x){
    fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)}
  )
  
  pred <- do.call("rbind", lapply(names(fit), function(x){
    do.call("rbind", lapply(names(fit[[x]]), function(y)
    {
      res <- data.frame(
        model = paste0("Model ", i),
        sex = x,
        age_bk = y,
        temp = fit[[x]][[y]]$pred$predvar,
        rr = fit[[x]][[y]]$pred$allRRfit,
        rrlow = fit[[x]][[y]]$pred$allRRlow,
        rrhigh = fit[[x]][[y]]$pred$allRRhigh
      )
    }))
  }))
  return(pred)
})

rr_curve_sensi <- do.call("rbind", rr_curve_sensi)
rr_curve_sensi$model <- as.factor(rr_curve_sensi$model)
rr_curve_sensi$model <-  factor(rr_curve_sensi$model, levels = paste0("Model ", 1:10))

# Create RR plot for women
RRplot_f <- filter(rr_curve_sensi, sex == "f") %>%
  ggplot(aes(x = temp, y = rr)) +
  geom_line(aes(color = model)) +
  geom_hline(
    yintercept = 1,
    linetype = "solid", color = "grey40"
  ) +
  facet_wrap(~ age_bk, ncol = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",    
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA)) +
  ylim(c(0.6, 3)) +
  scale_color_manual(values = brewer.pal(n = 10, name = "Set3")) +
  labs(y = "Relative Risk", x = "Temperature [C]", color = "")

# Create RR plot for men
RRplot_m <- filter(rr_curve_sensi, sex == "m") %>%
  ggplot(aes(x = temp, y = rr)) +
  geom_line(aes(color = model)) +
  geom_hline(
    yintercept = 1,
    linetype = "solid", color = "grey40"
  ) +
  facet_wrap(~ age_bk, ncol = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",    
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA)) +
  ylim(c(0.6, 3)) +
  scale_color_manual(values = brewer.pal(n = 10, name = "Set3")) +
  labs(y = "Relative Risk", x = "Temperature [C]", color = "")
```

This code generates Figure 22 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F, comment= F}

# Display figure for women
RRplot_f

# save figure
sauveFigure(config = 7,
            title = paste0("fig22"),
            path = folderFigures)
RRplot_f
dev.off()
```

This code generates Figure 23 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F, comment= F}

# Display figure for men
RRplot_m

# save figure
sauveFigure(config = 7,
            title = paste0("fig23"),
            path = folderFigures)
RRplot_m
dev.off()
```

## Sensitity to the lag hyperparameter

In this section, we compute the response
curves in terms of cumulative RR for women and men, using lags of 7, 14,
and 21 days. 

```{r, echo=TRUE, fig.width=12, fig.height=8 , message=F, warning=F, comment= F}
## Compar RR curve
rr_curve_sensi <- lapply(c(7, 14, 21), function(lagx){
  
  #Parameter
  param_dlnm_whole <- list(
    # main model, cubic natural spline with three internal knots in
    #   the 10th, 75th, 90th percentiles of the temperature distribution
    varfun = "bs",
    vardegree = 2,
    varper  = c(10, 75, 90),
    ## specification of the lag function
    # Definition of the maximum lag, that is, 21 days
    lag  = lagx,
    lagnk  = 3,
    ## degree of freedom for seasonality
    dfseas  = 8,
    ## degree of freedom for trend
    dftrend = NULL
  )
  
  fit <- lapply(base_nb_deces, function(x){
    fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)}
  )
  
  pred <- do.call("rbind", lapply(names(fit), function(x){
    do.call("rbind", lapply(names(fit[[x]]), function(y)
    {
      res <- data.frame(
        model = paste0("Lag ", lagx),
        sex = x,
        age_bk = y,
        temp = fit[[x]][[y]]$pred$predvar,
        rr = fit[[x]][[y]]$pred$allRRfit,
        rrlow = fit[[x]][[y]]$pred$allRRlow,
        rrhigh = fit[[x]][[y]]$pred$allRRhigh
      )
    }))
  }))
  
  stat <- do.call("rbind", lapply(names(fit), function(x){
    do.call("rbind", lapply(names(fit[[x]]), function(y)
    {
      data.frame(
        model = paste0("Lag ", lagx),
        sex = x,
        age_bk = y,
        qaic = fqaic(fit[[x]][[y]]$model),
        qbic = fqbic(fit[[x]][[y]]$model)
      )
    }))
  }))
  
  return(list(pred = pred, stat = stat))
})

# Manage results in a data frame
tab_stat <- do.call("rbind", lapply(rr_curve_sensi, function(rr){rr$stat}))
rr_curve_sensi <- do.call("rbind", lapply(rr_curve_sensi, function(rr){rr$pred}))
rr_curve_sensi$model <- as.factor(rr_curve_sensi$model)
rr_curve_sensi$model <-  factor(rr_curve_sensi$model, 
                                levels = paste0("Lag ", c(7, 14, 21)))

# Create RR plot for women
RRplot_f <- filter(rr_curve_sensi, sex == "f") %>%
  ggplot(aes(x = temp, y = rr)) +
  geom_line(aes(color = model)) +
  geom_hline(
    yintercept = 1,
    linetype = "solid", color = "grey40"
  ) +
  facet_wrap(~ age_bk, ncol = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",    
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA)) +
  ylim(c(0.6, 3)) +
  scale_color_manual(values = brewer.pal(n = 10, name = "Set3")) +
  labs(y = "Relative Risk", x = "Temperature [C]", color = "")

# Create RR plot for men
RRplot_m <- filter(rr_curve_sensi, sex == "m") %>%
  ggplot(aes(x = temp, y = rr)) +
  geom_line(aes(color = model)) +
  geom_hline(
    yintercept = 1,
    linetype = "solid", color = "grey40"
  ) +
  facet_wrap(~ age_bk, ncol = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",    
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA)) +
  ylim(c(0.6, 3)) +
  scale_color_manual(values = brewer.pal(n = 10, name = "Set3")) +
  labs(y = "Relative Risk", x = "Temperature [C]", color = "")
```

This code generates Figure 24 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8}

# Display figure for women
RRplot_f

# save figure
sauveFigure(config = 7,
            title = paste0("fig24"),
            path = folderFigures)
RRplot_f
dev.off()
```


This code generates Figure 25 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8 }

# Display figure for men
RRplot_m

# save figure
sauveFigure(config = 7,
            title = paste0("fig25"),
            path = folderFigures)
RRplot_m
dev.off()
```

## Sensitity to the calibration span

```{r, echo=TRUE, fig.width=12, fig.height=8}

## Compar RR curve
period_sensi <- c(2011, 2013, 2015, 2017, 2019)
rr_curve_sensi <- lapply(period_sensi, function(endx){
  
  #Parameter
  param_dlnm_whole <- list(
    # main model, cubic natural spline with three internal knots in
    #   the 10th, 75th, 90th percentiles of the temperature distribution
    varfun = "bs",
    vardegree = 2,
    varper  = c(10, 75, 90),
    ## specification of the lag function
    # Definition of the maximum lag, that is, 21 days
    lag  = 21,
    lagnk  = 3,
    ## degree of freedom for seasonality
    dfseas  = 8,
    ## degree of freedom for trend
    dftrend = NULL
  )
  
  fit <- lapply(base_nb_deces, function(x){
    # Filter on the YEARS
    x <- filter(x, YEARS %in% annee_deb:endx)
    fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)}
  )
  
  pred <- do.call("rbind", lapply(names(fit), function(x){
    do.call("rbind", lapply(names(fit[[x]]), function(y)
    {
      res <- data.frame(
        model = paste0("Period 1980-", endx),
        sex = x,
        age_bk = y,
        temp = fit[[x]][[y]]$pred$predvar,
        rr = fit[[x]][[y]]$pred$allRRfit,
        rrlow = fit[[x]][[y]]$pred$allRRlow,
        rrhigh = fit[[x]][[y]]$pred$allRRhigh
      )
    }))
  }))
  
  stat <- do.call("rbind", lapply(names(fit), function(x){
    do.call("rbind", lapply(names(fit[[x]]), function(y)
    {
      data.frame(
        model = paste0("Period 1980-", endx),
        sex = x,
        age_bk = y,
        qaic = fqaic(fit[[x]][[y]]$model),
        qbic = fqbic(fit[[x]][[y]]$model)
      )
    }))
  }))
  
  return(list(pred = pred, stat = stat))
})

# Manage results in a data frame
tab_stat <- do.call("rbind", lapply(rr_curve_sensi, function(rr){rr$stat}))
rr_curve_sensi <- do.call("rbind", lapply(rr_curve_sensi, function(rr){rr$pred}))
rr_curve_sensi$model <- as.factor(rr_curve_sensi$model)
rr_curve_sensi$model <-  factor(rr_curve_sensi$model, levels = paste0("Period 1980-", period_sensi))

# Create RR plot for women
RRplot_f <- filter(rr_curve_sensi, sex == "f") %>%
  ggplot(aes(x = temp, y = rr)) +
  geom_line(aes(color = model)) +
  geom_hline(
    yintercept = 1,
    linetype = "solid", color = "grey40"
  ) +
  facet_wrap(~ age_bk, ncol = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",    
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA)) +
  ylim(c(0.6, 3)) +
  scale_color_manual(values = brewer.pal(n = 10, name = "Set3")) +
  labs(y = "Relative Risk", x = "Temperature [C]", color = "")

# Create RR plot for men
RRplot_m <- filter(rr_curve_sensi, sex == "m") %>%
  ggplot(aes(x = temp, y = rr)) +
  geom_line(aes(color = model)) +
  geom_hline(
    yintercept = 1,
    linetype = "solid", color = "grey40"
  ) +
  facet_wrap(~ age_bk, ncol = 2) +
  theme_bw() +
  theme(
    legend.position = "bottom",    
    strip.background = element_rect(fill = "white", color = NA),
    strip.text = element_text(color = "black"),
    panel.border = element_rect(color = "black", fill = NA)) +
  ylim(c(0.6, 3)) +
  scale_color_manual(values = brewer.pal(n = 10, name = "Set3")) +
  labs(y = "Relative Risk", x = "Temperature [C]", color = "")
```


This code generates Figure 26 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8 }

# Display figure for women
RRplot_f

# save figure
sauveFigure(config = 7,
            title = paste0("fig26"),
            path = folderFigures)
RRplot_f
dev.off()
```

This code generates Figure 27 of the paper.

```{r, echo=TRUE, fig.width=12, fig.height=8 }

# Display figure for men
RRplot_m

# save figure
sauveFigure(config = 7,
            title = paste0("fig27"),
            path = folderFigures)
RRplot_m
dev.off()
```

# Session Information

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```
