---
title: "Results presented in Section 6.3 and in Appendix D"
author: "Quentin Guibert"
date: "2025-06-23"
institute: "Université Paris-Dauphine"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, fig.retina = 2)
options(encoding = 'UTF-8')
Sys.setlocale("LC_TIME", "English")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Folds
folder <- getwd() # If necessary, please set your working directory
fold_data <- paste0(folder, "/data/")
fold_data_raw <- paste0(folder, "/data_raw/")
fold_bib <- paste0(folder, "/functions/")
folderTables <- paste0(folder, "/tables/")
folderFigures <- paste0(folder, "/figures/")
#---------------------------------------------------------------------------------------------------
#### Load functions : ####
# Source functions
for (f in list.files(fold_bib))
  source(paste(fold_bib,f,sep=""), encoding = "UTF-8")

####  Load packages : ####
invisible(lapply(c(
  "MultiMoMo",
  "HMDHFDplus",
  "lattice",
  "grid",
  "ggplot2",
  "gridExtra",
  "locfit",
  "scales",
  "lubridate",
  "splines",
  "mgcv",
  "dlnm",
  "MASS",
  "mvtnorm",
  "Rfast",
  "parallel",
  "data.table",
  "formattable",
  "RColorBrewer",
  "readxl",
  "kableExtra",
  "dplyr",
  "tidyr",
  "tidyverse",
  "rmarkdown",
  "ggthemes",
  "cowplot",
  "ggridges",
  "viridis",
  "hrbrthemes",
  "colorspace",
  "ggbeeswarm",
  "purrr",
  "ggfan"),
  instal.import.package))

#----------------------------------------------------------
#theme graphique
theme_set(theme_bw())
trellis.device(color = FALSE)
```

# Parameters definition

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Main parameters
list_sexe <- list(m = 1,f = 2)
age_breaks <- c(0, 64, 74, 84, Inf)
age_labels <- c("0-64", "65-74", "75-84", "85+")

# training period
annee_deb <- 1980
annee_fin <- 2019
outlier <- 999 # exclude no atypical year

# forecasting period
start_year <- 2020
end_year <- 2100
year_breaks <- c(1980, 1989, 1999, 2009, 2019, 2029, 2039, 2049, 2059, 2069, 2079, 2089, 2100)
year_labels <- c("1980s", "1990s", "2000s", "2010s", "2020s", "2030s", "2040s",
                 "2050s", "2060s", "2070s", "2080s","2090s")

config_temp <-c(
  "all_effect", "hot_effect", "cold_effect", "moderate_hot_effect",
  "moderate_cold_effect","extr_hot_effect", "extr_cold_effect", "canicule_effect"
)

# simulation
nsim <- 10 # number of simulations
parallel <- "snow" # parallel option
ncpus <- 4 # number of cores

# other parameters
AGE_MAX <- 105
```

Here, we define the mortality model parameters

```{r, echo=TRUE, message=FALSE, warning=FALSE}

# Parameters for the Li-Lee model

xv <- 0:105
yv <- 1980:2019
group <- c("Female", "Male", "Total")

arima_spec <- list(K.t_M = "RWD", k.t_M = "AR1.1", k.t_F = "AR1.1")
n_ahead    <- length(start_year:end_year)
n_sim      <- 1000
est_method <- "PORT"
```

# Load data series and fit the DLNM model

## Load data

We import daily data series for death counts and temperatures. In the following lines,
we prepare our data before fiting the DLNM model. 

```{r, echo=TRUE}
base_nb_deces <- lapply(list_sexe, function(s){
  import_data_death(s, annee_deb, annee_fin, outlier)
})

base_nb_deces <- lapply(base_nb_deces, function(data){
  data$age_bk <- ifelse(
    data$age_bk == "0-49", "0-64", ifelse(
      data$age_bk == "50-54", "0-64", ifelse(
        data$age_bk == "55-59", "0-64", ifelse(
          data$age_bk == "60-64", "0-64", ifelse(
            data$age_bk == "65-69", "65-74", ifelse(
              data$age_bk == "70-74", "65-74", ifelse(
                data$age_bk == "75-79", "75-84", ifelse(
                  data$age_bk == "80-84", "75-84", ifelse(
                    data$age_bk == "85-89", "85+", ifelse(
                      data$age_bk == "90 et +", "85+", NA
                    ))))))))))
  
  data <- data %>%
    group_by(DateDec, age_bk) %>%
    summarise(Nombre_de_deces = sum(Nombre_de_deces),
              YEARS = min(YEARS),
              TAVG = min(TAVG),
              Ind_canicule = min(Ind_canicule),
              time = min(time),
              dow = min(dow),
              month = min(month)
    )
  return(as.data.table(data))
})
# Save results
save(base_nb_deces, file =  paste0(fold_data, "/base_nb_deces.RData"))
```

## Fit the DLNM model

We fit DLNM models for each gender and each age bucket. These models are calibrated
over all the years. 

```{r, echo=TRUE, fig.width=12, fig.height=8 }
# DLNM parameters
param_dlnm_whole <- list(
  # main model, cubic natural spline with three internal knots in
  #   the 10th, 75th, 90th percentiles of the temperature distribution
  varfun = "bs",
  vardegree = 2,
  varper  = c(10,75,90),
  ## specification of the lag function
  # Definition of the maximum lag, that is, 21 days
  lag  = 21,
  lagnk  = 3,
  ## degree of freedom for seasonality
  dfseas  = 8,
  ## degree of freedom for trend
  dftrend = NULL
)
```

## Model estimation
```{r echo=TRUE, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
fit <- lapply(base_nb_deces, function(x){
  fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)}
)
save(fit, file =  paste0(fold_data, "fit_dlnm.RData"))
```

## Estimation of the excess mortality for past periods

In order to estimate the death counts attributable to temperature for past periods, we compute the
average total attributable fraction.

### Calculating excess of mortality by year
```{r, echo=TRUE, fig.width=12, fig.height=8 }
# Quantiles for extreme hot and cold
q025 <- quantile(base_nb_deces$m$TAVG, 0.025)
q975 <- quantile(base_nb_deces$m$TAVG, 0.975)

# Run Excess of mortality
xs_mort <- lapply(names(base_nb_deces), function(x){
  res <- excess_mort_dlnm(base_nb_deces[[x]], fit[[x]], q_range = c(q025, q975),
                          nsim = 1, 
                          parallel = "no", ncpus = ncpus)
  return(res)
})
names(xs_mort) <- names(base_nb_deces)

# Tidy data into a data frame
xs_data <- lapply(names(xs_mort), function(i){
  xs_mort[[i]]$excess_mort_year %>%
    mutate(gender = i) %>%
    rename(year = YEARS)
})
xs_data <- do.call("rbind", xs_data)

# Save results
save(xs_mort, file =  paste0(fold_data, "xs_mort_central.RData"))
save(xs_data, file =  paste0(fold_data,  "xs_data.RData"))
```


### Extracting attributable fraction 

We select `all_effect` and consider that the effect is constant for all
ages in each bucket.

```{r, echo=TRUE, fig.width=12, fig.height=8}
# Data for women and men
xs_f <- xs_data %>% 
  dplyr::filter(gender == "f", temp_effect  == "all_effect") %>%
  dplyr::select(year, age_bk, ajust_factor)

xs_m <- xs_data %>% 
  dplyr::filter(gender == "m", temp_effect  == "all_effect") %>%
  dplyr::select(year, age_bk, ajust_factor)

# Transformation to matrix
transf_attrib_matrix <- function(df){
  mat <- matrix(data = 1, ncol = length(xv), nrow = length(yv))
  for(tt in yv){
    mat[tt - yv[1] + 1, 1:65] <- filter(df, year == tt, age_bk == "0-64")$ajust_factor
    mat[tt - yv[1] + 1, 66:75] <- filter(df, year == tt, age_bk == "65-74")$ajust_factor
    mat[tt - yv[1] + 1, 76:85] <- filter(df, year == tt, age_bk == "75-84")$ajust_factor
    mat[tt - yv[1] + 1, 86:(max(xv) + 1)] <- filter(df, year == tt, age_bk == "85+")$ajust_factor
  }
  dimnames(mat) <- list(yv, xv)
  return(mat)
}

xs_f <- transf_attrib_matrix(xs_f)
xs_m <- transf_attrib_matrix(xs_m)
```

# Li-Lee model

## Loading the mortality data

```{r, echo=TRUE, fig.width=12, fig.height=8}

# Get death and exposure data
load(file =  paste0(fold_data, "/dtx_list.RData"))
load(file =  paste0(fold_data, "/etx_list.RData"))

# Create datasets for the original Li-Lee model
mort_dt <- list(
  UNI = list(
    f = list(dtx = dtx_list$Female,
             etx = etx_list$Female,
             wa = w_matrix(dtx_list$Female)),
    m = list(dtx = dtx_list$Male,
             etx = etx_list$Male,
             wa = w_matrix(dtx_list$Male))),
  ALL = list(dtx = dtx_list$Total,
             etx = etx_list$Total,
             wa = w_matrix(dtx_list$Total))
)

# Create datasets for the Li-Lee model with adjuted exposure to risk
adj_mort_dt <- list(
  UNI = list(
    f = list(dtx = round(dtx_list$Female),
             etx = etx_list$Female * xs_f,
             wa = w_matrix(dtx_list$Female)),
    m = list(dtx = round(dtx_list$Male),
             etx = etx_list$Male * xs_m,
             wa = w_matrix(dtx_list$Male))),
  ALL = list(dtx = dtx_list$Total,
             etx = etx_list$Total,
             wa = w_matrix(dtx_list$Total))
)
## replace exposure to adjusted exposure
adj_mort_dt$ALL$etx <- adj_mort_dt$UNI$f$etx + adj_mort_dt$UNI$m$etx

## Save data
save(mort_dt, file =  paste0(fold_data, "mort_dt.RData"))
save(adj_mort_dt, file =  paste0(fold_data, "adj_mort_dt.RData"))
```


## Calibrating the Li-Lee models

We calibrate the both Li-Lee models with observed deaths and adjusted exposure to risk.

```{r, echo=TRUE, fig.width=12, fig.height=8}
# Original Li-Lee model
fit_M <- fit_li_lee(xv, yv, yv, "m", data = mort_dt, method = "NR", Ax = TRUE, exclude_coh = FALSE)
fit_F <- fit_li_lee(xv, yv, yv, "f", data = mort_dt, method = "NR", Ax = TRUE, exclude_coh = FALSE)

save(fit_M, file =  paste0(fold_data, "fit_M.RData"))
save(fit_F, file =  paste0(fold_data, "fit_F.RData"))
```

```{r, echo=TRUE, fig.width=12, fig.height=8}
# Li-Lee model with adjusted exposure to risk
adj_fit_M <- fit_li_lee(xv, yv, yv, "m", data = adj_mort_dt, method = "NR", Ax = TRUE, exclude_coh = FALSE)
adj_fit_F <- fit_li_lee(xv, yv, yv, "f", data = adj_mort_dt, method = "NR", Ax = TRUE, exclude_coh = FALSE)

save(adj_fit_M, file =  paste0(fold_data, "adj_fit_M.RData"))
save(adj_fit_F, file =  paste0(fold_data, "adj_fit_F.RData"))
```

## Running projection

```{r}
set.seed(1234)
proj_par <- project_parameters(fit_M, fit_F, n_ahead, n_sim, arima_spec, est_method)
adj_proj_par <- project_parameters(adj_fit_M, adj_fit_F, n_ahead, n_sim, arima_spec, est_method)
```

# Results of Section 6.3

## Estimated parameters of the Li-Lee model

This code generates Figure 7 of the paper. 

```{r, fig.width=12, fig.height=12}
list_fit <- list(fit_F = fit_F,
                 fit_M = fit_M,
                 adj_fit_F = adj_fit_F,
                 adj_fit_M = adj_fit_M)

# Display figure
myplot_parameters_li_lee(xv, yv, list_fit)

# save figure
sauveFigure(config = 7,
            title = paste0("fig7"),
            path = folderFigures)
myplot_parameters_li_lee(xv, yv, list_fit)
dev.off()
```

## Pearson residuals of the Li-Lee model

This code generates Figure 8 of the paper. 

```{r, fig.width=12, fig.height=4}
fig_res_f <- plot_residuals_li_lee(xv, yv, adj_fit_F, "FR", "Female")
fig_res_m <- plot_residuals_li_lee(xv, yv, adj_fit_M, "FR", "Male")

# Display figure
plot_grid(fig_res_f + ggtitle("Females"), fig_res_m + ggtitle('Males'),
          label_size = 12)

# save figure
sauveFigure(config = 9,
            title = paste0("fig8"),
            path = folderFigures)
plot_grid(fig_res_f + ggtitle("Females"), fig_res_m + ggtitle('Males'),
          label_size = 12)
dev.off()
```

## Calibrated parameters in the Li-Lee models

This code generates Table 2 of the paper. 

```{r}

# Create table
tab_param <- data.frame(
  v1 = c("Original Li-Lee model", "Adjusted exposure to risk"),
  v2 = c(proj_par$coef_KtM, adj_proj_par$coef_KtM),
  v3 = c(proj_par$coef_ktM[1], adj_proj_par$coef_ktM[1]),
  v4 = c(proj_par$coef_ktM[2], adj_proj_par$coef_ktM[2]),
  v5 = c(proj_par$coef_ktF[1], adj_proj_par$coef_ktF[1]),
  v6 = c(proj_par$coef_ktF[2], adj_proj_par$coef_ktF[2])
)

# Print Table
tab_param %>%
  kbl(booktabs = T, 
      align='c',
      label = "li_lee_parameters",
      col.names = c("Calibration data", "$\\delta$",
                    "$c^{(m)}$", "$\\phi^{(m)}$",
                    "$c^{(f)}$", "$\\phi^{(f)}$"),
      digits=4, 
      escape =  FALSE) %>%
  kable_classic_2(full_width = F)

# Save table in .tex
tab_param %>%
  kbl(booktabs = T, format = 'latex', 
      align='c',
      label = "li_lee_parameters",
      col.names = c("Calibration data", "$\\delta$",
                    "$c^{(m)}$", "$\\phi^{(m)}$",
                    "$c^{(f)}$", "$\\phi^{(f)}$"),
      digits=4, 
      escape =  FALSE) %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(latex_options = "hold_position",  font_size = 8) %>%
  readr::write_lines(file = paste0(folderTables, "tab2.tex"))
```


## Correlation matrix of residuals in the Li-Lee models

This code generates Table 3 of the paper. 

```{r}
# Correction matrix function
corr_function <- function(S)
{
  D <- diag(sqrt(diag(S)))
  Dinv <- solve(D)
  R <- Dinv %*% S %*% Dinv
  dimnames(R) <- dimnames(S)
  R 
}

# Create Correction matrix into a table
tab_corr <- cbind(data.frame(corr_function(proj_par$cov_mat)), 
                  data.frame(corr_function(adj_proj_par$cov_mat)))
names(tab_corr) <- paste0("v", 1:ncol(tab_corr))
row.names(tab_corr) <- NULL

# Print Table
tab_corr %>%
  dplyr::mutate(v = c("$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$"), .before = "v1") %>%
  kbl(booktabs = T, 
      align='c',
      label = "li_lee_corrmatrix",
      col.names = c("", "$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$",
                    "$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$"),
      digits=4, 
      escape =  FALSE) %>%
  add_header_above(c(" " = 1, "Original Li-Lee model" = 3, "Adjusted exposure to risk" = 3)) %>%
  kable_classic_2(full_width = F)

# Save table in .tex
tab_corr %>%
  dplyr::mutate(v = c("$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$"), .before = "v1") %>%
  kbl(booktabs = T, format = 'latex', 
      align='c',
      label = "li_lee_corrmatrix",
      col.names = c("", "$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$",
                    "$e_t$", "$r_t^{(m)}$", "$r_t^{(f)}$"),
      digits=4, 
      escape =  FALSE) %>%
  add_header_above(c(" " = 1, "Original Li-Lee model" = 3, "Adjusted exposure to risk" = 3)) %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(latex_options = "hold_position",  font_size = 8) %>%
  readr::write_lines(file = paste0(folderTables, "tab3.tex"))
```


## Projection of trend parameters

This code generates Figure 9 of the paper. 

```{r, fig.width=12, fig.height=4}
# Projection of time dependant parameters
p <- plot_kt_sim(proj_par, adj_proj_par, n_sim, yv[1], end_year)

# Display figure
print(p)

# Save figure
sauveFigure(config = 9,
            title = paste0("fig9"),
            path = folderFigures)
print(p)
dev.off()
```

# Results of Appendix D.1

## Comparison between observed rates and predicted rates

```{r, fig.width=12, fig.height=4}
## project mortality rates
proj_rates <- project_mortality_rates(fit_M, fit_F, proj_par)
adj_proj_rates <- project_mortality_rates(adj_fit_M, adj_fit_F, adj_proj_par)

## Function to extract median rates
format_rates_median <- function(rates)
{
  rates <- melt(rates)
  names(rates) <- c("age", "year", "sim", "Qxt")
  rates <- rates %>% 
    mutate(age_bk = cut(age, age_breaks, age_labels, include.lowest = T),
           mxt = -log(1 - Qxt)) %>%
    group_by(age, year, age_bk) %>%
    summarise(mxt = median(mxt)) %>%
    filter(year < start_year, age %in% c(45, 65, 75, 85))
}

# Project median mortality rates for the original Li-Lee model
proj_rates <- lapply(proj_rates, format_rates_median)
names(proj_rates) <- c("m", "f")

# Project median mortality rates for the Li-Lee model with adjuted exposure to risk
adj_proj_rates <- lapply(adj_proj_rates, format_rates_median)
names(adj_proj_rates) <- c("m", "f")
```

In the following chuncks, we compare : 

1.    the observed rates
2.    the predicted rates with the original LL model
3.    the predicted rates with Li-Lee model when the calibration process used
adjusted exposure. The adjustment factor is estimated based observed deaths
and temperatures
4.   the predicted rates with Li-Lee model when the calibration process used
adjusted exposure. The adjustment factor is predicted based on temperature only. 
We use no weight and weight calibrated on the last five years.

We focus on the prediction listed in 4. 

```{r, fig.width=12, fig.height=4}
load(file =  paste0(fold_data, "/xs_mort_central.RData"))

# Computing annual weight
un_an_weight <- lapply(names(xs_mort), function(x){
  # Add day and months
  # Filter on the last five years
  res <- xs_mort[[x]]$excess_mort %>%
    mutate(day = day(DateDec),
           month = month(DateDec)) %>%
    filter(YEARS %in% c(2015:2019), temp_effect == "all_effect") %>%
    mutate(un_an = cases - an) %>%
    dplyr::select(YEARS, age_bk, un_an, day, month) %>%
    group_by(age_bk, day, month) %>%
    summarise(un_an = sum(un_an))
  
  # Calculate the denominator of the weight
  xs_mort_agg <- res %>%
    group_by(age_bk) %>%
    summarise(un_an_sum = sum(un_an))
  
  res <- res %>%
    left_join(xs_mort_agg, by = "age_bk") %>%
    mutate(weight = un_an / un_an_sum) %>%
    mutate(un_an = NULL,
           un_an_sum = NULL)
  # Correct 29-02
  res$weight[res$day==29 & res$month==2] <- 5 * 
    res$weight[res$day==29 & res$month==2]
  
  return(res)
})
names(un_an_weight) <- names(xs_mort)

# Computing mortality rates
xs_pred_4 <- lapply(names(xs_mort), function(x){
  temp <- xs_mort[[x]]$excess_mort %>%
    filter(temp_effect == "all_effect") %>%
    mutate(day = day(DateDec), month = month(DateDec)) %>%
    left_join(un_an_weight[[x]], by = join_by(day, month, age_bk)) %>%
    mutate(day = NULL, month = NULL) %>%
    mutate(af_weight = af * (1 - af)^(-1) * weight,
           af_no_weight = af * (1 - af)^(-1)) %>%
    group_by(YEARS, age_bk) %>%
    summarise(
      af_weight_year = sum(af_weight),
      af_no_weight_year = sum(af_no_weight),
      nb_days = n()) %>%
    mutate(af_no_weight_year = af_no_weight_year / nb_days) %>%
    mutate(af_no_weight_year = 1+af_no_weight_year,
           af_weight_year = 1 + af_weight_year,
           year = YEARS) 
  
})
names(xs_pred_4) <- names(xs_mort)
```

We now create a table for comparing results.

```{r, fig.width=12, fig.height=4}

tab_result <- lapply(names(proj_rates), function(x){
  
  #1. Obs rates
  obs_rates <- melt( mort_dt$UNI[[x]]$dtx)
  names(obs_rates) <- c("year", "age", "mxt")
  obs_rates$ext <- melt( mort_dt$UNI[[x]]$etx)[,3]
  obs_rates$mxt <-  obs_rates$mxt / obs_rates$ext
  obs_rates$ext <- NULL
  obs_rates$model = "Obs"
  
  #2. Pred rates original LL
  pred1 <- proj_rates[[x]] %>%
    mutate(model = "Original LL")
  
  #3. Pred rates LL with adjusted exp
  if(x == "m"){
    xs_gender <- xs_m
  } else {
    xs_gender <- xs_f
  }
  xs_melt <- melt(xs_gender)
  names(xs_melt) <- c("year", "age", "xs")
  xs_melt <- xs_melt %>%
    mutate(age_bk = cut(age, age_breaks, age_labels, include.lowest = T))
  
  pred2 <- adj_proj_rates[[x]] %>%
    mutate(model = "LL adj. (1)") %>%
    left_join(xs_melt, by = join_by(age, year)) %>%
    mutate(mxt = mxt * xs)
  
  #4. Pred rates LL with adjusted exp and predicted adjustement coef
  pred3 <- adj_proj_rates[[x]] %>%
    mutate(model = "LL adj. (2)") %>%
    left_join(xs_pred_4[[x]], by = join_by(age_bk, year)) %>%
    mutate(mxt = mxt * af_weight_year)
  
  pred4 <- adj_proj_rates[[x]] %>%
    mutate(model = "LL adj. (3)") %>%
    left_join(xs_pred_4[[x]], by = join_by(age_bk, year)) %>%
    mutate(mxt = mxt * af_no_weight_year)
  
  res <- rbind(obs_rates[, c("year", "age", "mxt", "model")],
               pred1[, c("year", "age", "mxt", "model")],
               pred2[, c("year", "age", "mxt", "model")],
               pred3[, c("year", "age", "mxt", "model")],
               pred4[, c("year", "age", "mxt", "model")])
  
  res$model <- factor(res$model, levels = c("Obs", "Original LL", "LL adj. (1)", "LL adj. (2)", "LL adj. (3)"))
  return(res)
})
names(tab_result) <- names(xs_mort)
```

## Comparison of observed and predicted central death rates for females

This code generates Figure 28 of the paper. 

```{r, fig.width=12, fig.height=12, eval = TRUE}
pf <- tab_result$f %>%
  filter(age %in% c(45, 65, 75, 85)) %>%
  ggplot(aes(x = year, y = mxt)) +
  geom_line(aes(colour = model), size = 0.8) +
  facet_wrap( ~ factor(age), scales = "free_y", ncol = 2) +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = "white", color = "black"),
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))+
  labs(x = "Years", y = "Central death rates", colour = "") +
  scale_color_manual(values = brewer.pal(n = 5, name = "Set1"))

# Display figure
pf

# Save figure
sauveFigure(config = 7,
            title = paste0("fig28"),
            path = folderFigures)
print(pf)
dev.off()
```


## Comparison of observed and predicted central death rates for males

This code generates Figure 29 of the paper. 

```{r, fig.width=12, fig.height=12}
pm <- tab_result$m %>%
  filter(age %in% c(45, 65, 75, 85)) %>%
  ggplot(aes(x = year, y = mxt)) +
  geom_line(aes(colour = model), size = 0.8) +
  facet_wrap( ~ factor(age), scales = "free_y", ncol = 2) +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = "white", color = "black"),
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))+
  labs(x = "Years", y = "Central death rates", colour = "") +
  scale_color_manual(values = brewer.pal(n = 5, name = "Set1"))

# Display figure
pm

# Save figure
sauveFigure(config = 7,
            title = paste0("fig29"),
            path = folderFigures)
print(pm)
dev.off()
```

# Results of Appendix D.2

In this section, we conduct an out-of-sample forecasting evaluation of our modeling framework. 
Building on the projection strategy described in Section 5, we apply a cross-validation procedure and
compare Li-Lee models estimated using the original exposure to risk and those adjusted for temperature.
The models are fitted to data from 1980 to 2011 (i.e. 80% of the sample) and evaluated on the 2012-2019 period (i.e. 20% of the sample).

We first modify parameters

```{r, fig.width=12, fig.height=4}
yv <- 1980:2011
# forecasting period
start_year <- 2012
end_year <- 2019

# Monte-Carlo simulations
nsim <- 1000L # number of simulations
# parallel <- "snow" # parallel option
parallel <- "no" # parallel option
ncpus <- 8L # number of cores

# other parameters
AGE_MAX <- 105

## Forecasting Parameters
arima_spec <- list(K.t_M = "RWD", k.t_M = "AR1.1", k.t_F = "AR1.1")
n_ahead    <- length(start_year:end_year)
n_sim      <- nsim
est_method <- "PORT"

q025 <- quantile(base_nb_deces$m$TAVG, 0.025)
q975 <- quantile(base_nb_deces$m$TAVG, 0.975)
```

## The DLNM model

### Fit the model on in-sample period

```{r, fig.width=12, fig.height=4}

# DLNM parameters
param_dlnm_whole <- list(
  # main model, cubic natural spline with three internal knots in
  #   the 10th, 75th, 90th percentiles of the temperature distribution
  varfun = "bs",
  vardegree = 2,
  varper  = c(10,75,90),
  ## specification of the lag function
  # Definition of the maximum lag, that is, 21 days
  lag  = 21,
  lagnk  = 3,
  ## degree of freedom for seasonality
  dfseas  = 8,
  ## degree of freedom for trend
  dftrend = NULL
)

# Fit the DLNM model on the in-sample period
fit_insample <- lapply(base_nb_deces, function(x){
  # Filter on insample period
  x <- filter(x, YEARS %in% yv)
  fit_dlnm(x, param_dlnm_whole, per_age = T, summer = F, psi = NULL)}
)
```

### Calculating excess mortality by year

```{r, echo=TRUE, fig.width=12, fig.height=8}
# Run Excess of mortality
xs_mort_insample <- lapply(names(base_nb_deces), function(x){
  # Filter on insample period
  temp_data <- filter(base_nb_deces[[x]], YEARS %in% yv)
  res <- excess_mort_dlnm(temp_data, fit_insample[[x]], 
                          q_range = c(q025, q975),
                          nsim = 1, 
                          parallel = "no", ncpus = ncpus)
  return(res)
})
names(xs_mort_insample) <- names(base_nb_deces)

xs_data_insample <- lapply(names(xs_mort_insample), function(i){
  xs_mort_insample[[i]]$excess_mort_year %>%
    mutate(gender = i) %>%
    rename(year = YEARS)
})
xs_data_insample <- do.call("rbind", xs_data_insample)
```

```{r, echo=TRUE, fig.width=12, fig.height=8}
# Data for women and men
xs_f_insample <- xs_data_insample %>% 
  dplyr::filter(gender == "f", temp_effect  == "all_effect") %>%
  dplyr::select(year, age_bk, ajust_factor)

xs_m_insample <- xs_data_insample %>% 
  dplyr::filter(gender == "m", temp_effect  == "all_effect") %>%
  dplyr::select(year, age_bk, ajust_factor)

# Transformation to matrix
xs_f_insample <- transf_attrib_matrix(xs_f_insample)
xs_m_insample <- transf_attrib_matrix(xs_m_insample)
```

We compute this weight as an average over the last five years.

```{r}
un_an_weight <- lapply(names(xs_mort_insample), function(x){
  # Add day and months
  # Filter on the last five years
  res <- xs_mort_insample[[x]]$excess_mort %>%
    mutate(day = day(DateDec),
           month = month(DateDec)) %>%
    filter(YEARS %in% c((start_year - 5):(start_year - 1)), temp_effect == "all_effect") %>%
    mutate(un_an = cases - an) %>%
    dplyr::select(YEARS, age_bk, un_an, day, month) %>%
    group_by(age_bk, day, month) %>%
    summarise(un_an = sum(un_an))
  
  # Calculate the denominator of the weight
  xs_mort_agg <- res %>%
    group_by(age_bk) %>%
    summarise(un_an_sum = sum(un_an))
  
  res <- res %>%
    left_join(xs_mort_agg, by = "age_bk") %>%
    mutate(weight = un_an / un_an_sum) %>%
    mutate(un_an = NULL,
           un_an_sum = NULL)
  # Correct 29-02
  res$weight[res$day==29 & res$month==2] <- 5 * 
    res$weight[res$day==29 & res$month==2]
  
  return(res)
})
names(un_an_weight) <- names(xs_mort_insample)
```

## Fit the Li-Lee model

```{r}
# Create datasets on the in-sample period
range_hist <- seq_len(length(yv))

# Create datasets for the LL model with original exposure
mort_dt_insample <- list(
  UNI = list(
    f = list(dtx = dtx_list$Female[range_hist,],
             etx = etx_list$Female[range_hist,],
             wa = w_matrix(dtx_list$Female[range_hist,])),
    m = list(dtx = dtx_list$Male[range_hist,],
             etx = etx_list$Male[range_hist,],
             wa = w_matrix(dtx_list$Male[range_hist,]))),
  ALL = list(dtx = dtx_list$Total[range_hist,],
             etx = etx_list$Total[range_hist,],
             wa = w_matrix(dtx_list$Total[range_hist,]))
)

# Create datasets for the LL model with ajusted exposure
adj_mort_dt_insample <- list(
  UNI = list(
    f = list(dtx = round(dtx_list$Female[range_hist,]),
             etx = (etx_list$Female[range_hist,] * xs_f_insample),
             wa = w_matrix(dtx_list$Female[range_hist,])),
    m = list(dtx = round(dtx_list$Male[range_hist,]),
             etx = (etx_list$Male[range_hist,] * xs_m_insample),
             wa = w_matrix(dtx_list$Male[range_hist,]))),
  ALL = list(dtx = dtx_list$Total[range_hist,],
             etx = etx_list$Total[range_hist,],
             wa = w_matrix(dtx_list$Total[range_hist,]))
)

## replace exposure to adjusted exposure
adj_mort_dt_insample$ALL$etx <- adj_mort_dt_insample$UNI$f$etx + adj_mort_dt_insample$UNI$m$etx
```

We calibrate the models both with observed deaths and adjusted exposure.
```{r}

# Original LL
fit_M_insample <- fit_li_lee(xv, yv, yv, "m", data = mort_dt_insample, method = "NR", Ax = TRUE, exclude_coh = FALSE)
fit_F_insample <- fit_li_lee(xv, yv, yv, "f", data = mort_dt_insample, method = "NR", Ax = TRUE, exclude_coh = FALSE)

# LL with adjusted exposure to risk
adj_fit_M_insample <- fit_li_lee(xv, yv, yv, "m", data = adj_mort_dt_insample, method = "NR", Ax = TRUE, exclude_coh = FALSE)
adj_fit_F_insample <- fit_li_lee(xv, yv, yv, "f", data = adj_mort_dt_insample, method = "NR", Ax = TRUE, exclude_coh = FALSE)
```

## Project Future mortality

We forecast the mortality rates with and without temperature effect. 

### Project mortality with the original LL model

```{r}

# project parameters
set.seed(1234)
proj_par_outsample <- project_parameters(fit_M_insample, fit_F_insample, n_ahead, n_sim, arima_spec, est_method)
set.seed(1234)
adj_proj_par_outsample <- project_parameters(adj_fit_M_insample, adj_fit_F_insample, n_ahead, n_sim, arima_spec, est_method)

## project mortality rates
proj_rates_outsample <- project_mortality_rates(fit_M_insample, fit_F_insample, proj_par_outsample)
adj_proj_rates_outsample <- project_mortality_rates(adj_fit_M_insample, adj_fit_F_insample, adj_proj_par_outsample)

# reformat data
format_rates <- function(rates)
{
  rates <- melt(rates)
  names(rates) <- c("age", "year", "sim", "Qxt")
  rates <- rates %>% 
    mutate(age_bk = cut(age, age_breaks, age_labels, include.lowest = T))
}

proj_rates_outsample <- lapply(proj_rates_outsample, format_rates)
names(proj_rates_outsample) <- c("m", "f")

adj_proj_rates_outsample <- lapply(adj_proj_rates_outsample, format_rates)
names(adj_proj_rates_outsample) <- c("m", "f")
```

### Project mortality with temperature effects

```{r, echo=TRUE, fig.width=12, fig.height=8 }
data_mort <- adj_proj_rates_outsample

# select temperature over the projection period
sc <- dplyr::filter(base_nb_deces$m, 
                    YEARS %in% start_year:end_year,
                    age_bk == "0-64") %>%
  dplyr::select(DateDec, YEARS, Ind_canicule, TAVG)

# run simulations for males and females
res <- lapply(names(data_mort), function(x)
{
  # select population
  temp_dem <- data_mort[[x]] %>%
    filter(year %in% start_year:end_year)
  # select weight
  temp_weight <- un_an_weight[[x]]
  
  # launch forecasting along the rcp scenario with simulations and parallelization
  f_mort <- forecast_mort_dlnm(temp_dem,
                               sc,
                               fit_insample[[x]],
                               temp_weight,
                               q_range = c(q025, q975),
                               age_qx = 0:105,
                               sensi_cen = 0,
                               nsim = nsim,
                               parallel = parallel, ncpus = ncpus, seed = 1234)
  # Add gender indexes
  f_mort <- lapply(f_mort, function(tt){
    if(is.null(tt))
    {
      return(tt)
    } else{
      return(
        tt %>% 
          dplyr::mutate(gender = x) %>%
          relocate(gender, temp_effect, .before =  sim) 
      )
    }
  })
  return(f_mort)
})

# Merge more than two lists with the same element name
res <- Reduce(function(...) Map("rbind", ...), res)

# Save
save(res, file = paste0(fold_data, "/out_of_sample_results.RData"))
```

## Result analysis

### Generating results

First, we collect simulations from the original mortality and compute life expectancy.

```{r}
# Compute life expectancy
real_rates <- proj_rates_outsample
real_rates$m$gender <- "m"
real_rates$f$gender <- "f"
real_rates <- do.call("rbind", real_rates)
real_rates <- real_rates%>%
  dplyr::filter(year %in% start_year:end_year)
real_rates <- life_exp_ts(real_rates, calc_xs = FALSE)

# Compute mxt
real_rates <- real_rates %>% 
  rename(Qxt_real = Qxt, Ext_real = Ext) %>%
  mutate(mxt_real = -log(1 - Qxt_real)) %>%
  dplyr::select(gender, sim, year, age, mxt_real, Ext_real)

# Aggregate and compute median
real_rates <- real_rates %>% 
  group_by(gender, year, age) %>%
  summarise(mxt = median(mxt_real),
            Ext = median(Ext_real)) %>%
  mutate(model = "Original LL")

# Create two tables
tab_Ext_original <- real_rates %>%
  filter(age ==0) %>%
  mutate(Ext_up = Ext,
         Ext_down = Ext) %>%
  dplyr::select(! mxt)

tab_mxt_original <- real_rates %>%
  mutate(mxt_up = mxt,
         mxt_down = mxt) %>%
  dplyr::select(! Ext)
```

We do the same with LL model with adjusted exposure to risk.

```{r}
# Life expectancy
tab_Ext_outsample <- res$tab_ex%>%
  filter(temp_effect == "all_effect") %>%
  dplyr::select(gender, sim, year, age, Ext_xs) %>%
  group_by(gender, year, age) %>%
  summarise(Ext_up = quantile(Ext_xs, 0.95),
            Ext_down = quantile(Ext_xs, 0.05),
            Ext = median(Ext_xs)) %>%
  mutate(model = "LL adj") %>%
  filter(age == 0)

# mxt 
tab_rates_outsample <- res$tab_qx %>%
  filter(temp_effect == "all_effect") %>%
  mutate(mxt = -log(1 - Qxt),
         mxt_xs = -log(1 - Qxt_xs)) %>%
  dplyr::select(gender, sim, year, age, mxt_xs) %>%
  group_by(gender, year, age) %>%
  summarise(mxt = median(mxt_xs),
            mxt_up = quantile(mxt_xs, 0.95),
            mxt_down = quantile(mxt_xs, 0.05)) %>%
  mutate(model = "LL adj")
```

Finally, we compute observed rates.

```{r}
# Obs rates
obs_rates <- do.call("rbind", 
                     lapply(names(mort_dt$UNI), function(x)
                     {
                       or <- melt(mort_dt$UNI[[x]]$dtx)
                       names(or) <- c("year", "age", "mxt")
                       or$ext <- melt(mort_dt$UNI[[x]]$etx)[,3]
                       or$mxt <-  or$mxt / or$ext
                       or$ext <- NULL
                       or$model <- "Obs"
                       or$gender = x
                       return(or)
                     }
                     )
)
```

### Comparison of observed and predicted log-mortality rates 

This code generates Figure 30 of the paper. 

```{r, fig.width=12, fig.height=6}
# Figure for log mortality rates at the tend of the observation period
g_final_rates <- obs_rates %>%
  mutate(mxt_up = mxt,
         mxt_down = mxt) %>%
  bind_rows(tab_rates_outsample) %>%
  bind_rows(tab_mxt_original) %>%
  filter(year == 2019) %>%
  mutate(model = factor(model, levels = c("Obs","Original LL","LL adj"))) %>%
  mutate(gender = fct_recode(gender, 
                             "Females" = "f",
                             "Males" = "m",
  )) %>%
  ggplot(aes(x = as.numeric(age), y = log(mxt))) +
  geom_line(aes(colour = factor(model))) +
  facet_wrap(~ gender, scales = "free_y", ncol = 2) +
  labs(x = "Ages", y = "Log(central death rates)", fill = "", colour = "") +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))

# Display figure
g_final_rates

# Save figure
sauveFigure(config = 8,
            title = paste0("fig30"),
            path = folderFigures)
print(g_final_rates)
dev.off()
```

### Comparison of RMSE of predicted log-mortality rates

First, we compute error metrics. 

```{r, fig.width=12, fig.height=6}

# Compute errors based on the log-mortality rates
compute_errors <- function(true_data, pred_data, model_name) {
  # Joint predicted and observed mxt
  df <- true_data %>%
    left_join(pred_data, by = c("age", "gender", "year"), suffix = c("_obs", "_pred")) %>%
    mutate(
      abs_err  = abs(log(mxt_obs) - log(mxt_pred)),
      sq_err   = (log(mxt_obs) - log(mxt_pred))^2,
      model    = model_name
    )
  
  # Group by  year
  by_year <- df %>%
    group_by(gender, year, model) %>%
    summarise(
      RMSE = sqrt(mean(sq_err)),
      MAE  = mean(abs_err),
      .groups = "drop"
    )
  
  # Group by  year
  by_age <- df %>%
    group_by(gender, age, model) %>%
    summarise(
      RMSE = sqrt(mean(sq_err)),
      MAE  = mean(abs_err),
      .groups = "drop"
    )
  
  # Aggregated results
  global <- df %>%
    group_by(gender, model) %>%
    summarise(
      RMSE = sqrt(mean(sq_err)),
      MAE  = mean(abs_err),
      .groups = "drop"
    ) %>%
    mutate(age = "ALL", year = "ALL")
  
  return(list(by_year = by_year,
              by_age = by_age,
              global = global))
}

# Apply for each gender
error_results <-  lapply(c("m", "f"), function(sex) 
{
  obs_sex <- obs_rates %>% 
    filter(gender == sex, year %in% start_year:end_year)
  
  r_original <- compute_errors(obs_sex,  tab_mxt_original %>% filter(gender == sex), model_name = "Original LL")
  r_outsample <- compute_errors(obs_sex,  tab_rates_outsample %>% filter(gender == sex), model_name = "LL adj")
  
  return(list(r_original = r_original, r_outsample = r_outsample))
})
names(error_results) <- c("m", "f")
```

This code generates Figure 31 of the paper. 

```{r, fig.width=12, fig.height=6}

# By year
g_rmse_year <- error_results$m$r_outsample$by_year %>%
  bind_rows(error_results$m$r_original$by_year,
            error_results$f$r_outsample$by_year,
            error_results$f$r_original$by_year) %>%
  mutate(model = factor(model, levels = c("Original LL","LL adj"))) %>%
  mutate(gender = fct_recode(gender, 
                             "Females" = "f",
                             "Males" = "m",
  )) %>%
  ggplot(aes(x = year, y = RMSE)) +
  geom_line(aes(colour = factor(model))) +
  facet_wrap(~ gender, scales = "free_y", ncol = 2) +
  labs(x = "Years", y = "RMSE", fill = "", colour = "") +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))

# Display figure
g_rmse_year

# Save figure
sauveFigure(config = 8,
            title = paste0("fig31"),
            path = folderFigures)
print(g_rmse_year)
dev.off()
```

This code generates Figure 32 of the paper. 

```{r, fig.width=12, fig.height=6}

# By age
g_rmse_age <- error_results$m$r_outsample$by_age %>%
  bind_rows(error_results$m$r_original$by_age,
            error_results$f$r_outsample$by_age,
            error_results$f$r_original$by_age) %>%
  mutate(model = factor(model, levels = c("Original LL","LL adj"))) %>%
  mutate(gender = fct_recode(gender, 
                             "Females" = "f",
                             "Males" = "m",
  )) %>%
  ggplot(aes(x = age, y = RMSE)) +
  geom_line(aes(colour = factor(model))) +
  facet_wrap(~ gender, scales = "free_y", ncol = 2) +
  labs(x = "Ages", y = "RMSE", fill = "", colour = "") +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = "white", color = NA),
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(color = "black"),
        panel.border = element_rect(color = "black", fill = NA),
        text = element_text(size = 12))

# Display figure
g_rmse_age

# Save figure
sauveFigure(config = 8,
            title = paste0("fig32"),
            path = folderFigures)
print(g_rmse_age)
dev.off()
```

This code generates Table 6 of the paper. 

```{r, fig.width=12, fig.height=4}

# Display table  of global error
error_results$m$r_original$global %>%
  bind_rows(error_results$m$r_outsample$global,
            error_results$f$r_original$global,
            error_results$f$r_outsample$global) %>%
  dplyr::select(gender, model, RMSE, MAE) %>%
  kbl( booktabs = T, 
       align='c',
       label = "out_of_sample_perf",
       col.names = c("Sex", "Model",
                     "RMSE", "MAE"),
       digits=4, 
       escape =  FALSE) %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(latex_options = "hold_position",  font_size = 8)

# Tab of global error
error_results$m$r_original$global %>%
  bind_rows(error_results$m$r_outsample$global,
            error_results$f$r_original$global,
            error_results$f$r_outsample$global) %>%
  dplyr::select(gender, model, RMSE, MAE) %>%
  kbl( booktabs = T, format = 'latex', 
       align='c',
       label = "out_of_sample_perf",
       col.names = c("Sex", "Model",
                     "RMSE", "MAE"),
       digits=4, 
       escape =  FALSE) %>%
  kable_classic_2(full_width = F) %>%
  kable_styling(latex_options = "hold_position",  font_size = 8) %>%
  readr::write_lines(file = paste0(folderTables, "tab6.tex"))
```


# Session Information

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sessionInfo()
```
